%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Clinical Information Extraction}
\label{chap:rw-ie}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Introduction}
\label{sec:ie-intro}

The previous chapters built pretrained language models and asked what data they should learn
from. This chapter turns to the task these models are built for. Clinical information
extraction means turning unstructured clinical text into structured information that a computer
can store, query, and analyse. A discharge summary, a referral letter, or a prescription is
written for human readers, so the facts it contains are not directly available to software. The
goal is to recover those facts, such as diagnoses, treatments, and their relations, and to
record them in a form that supports research and care.

This task matters because hospitals increasingly store their records in clinical data
warehouses, large databases that gather electronic health records for secondary use. Most of
the clinically useful information in these records lives in free text rather than in structured
fields. Most medical concepts appear only in the free text, which leaves reports, letters, and
prescriptions as the main source \citep{raghavan_how_2014}.
France adds a further reason to study the problem. The PMSI, the national system that describes
hospital activity, requires every hospital stay to be coded with ICD-10 diagnosis codes and
CCAM procedure codes, covering about 30 million stays a year. This coding is still done by hand
by professional coders and reported to the ATIH, the national agency for hospital information.
This makes automation a practical and economic priority.

Clinical information extraction covers a family of related tasks. Named entity recognition finds
the spans of text that mention entities of interest, such as a drug name or a symptom. Relation
extraction identifies how these entities are connected, for example which drug treats which
condition. Medical coding assigns ontology codes to a document, such as ICD-10, CCAM, or ATC
codes for medications. Entity linking, also called normalisation, maps a mention in the text to
a single concept in a knowledge base such as the UMLS or SNOMED-CT. These tasks are often
combined in a pipeline, since a coded record usually needs both the entities and the concepts
they refer to.

These tasks are the right place to close the related work, because they are where the
pretrained models of the earlier chapters are finally put to use and judged. Once a model has
been built and adapted, its value shows in how well it supports named entity recognition,
coding, and normalisation on real clinical text. The difficulties that recur across these tasks
also frame the open problems of the field. Labelled clinical data are scarce, because the text
is sensitive and annotation requires medical expertise. The label space can be very large,
since coding systems contain tens of thousands of codes. Evaluation is not standardised, which
makes results hard to compare across datasets and studies.

The chapter follows the structure of these tasks. It starts from the fundamentals of named
entity recognition, then reviews biomedical and clinical benchmarks. It moves on to medical
coding and to entity linking, before turning to zero-shot approaches and to the use of
synthetic data. It closes with a discussion of how clinical information extraction is
evaluated.
\section{Named Entity Recognition}
\label{sec:ie-ner}

Named entity recognition (NER) is the task of finding entity mentions in text and classifying
them into predefined types. It is often the first step in an extraction pipeline, since later
tasks such as relation extraction and entity linking build on the entities that NER returns.
The following subsections describe how the task has been formulated over time, from sequence
labeling with handcrafted models to pretrained encoders and span-based methods.

\subsection{Sequence Labeling}

The classic way to formulate NER is as a sequence labeling problem with the BIO tagging
scheme. Each token in the text receives one label: \texttt{B-TYPE} marks the beginning of an
entity of a given type, \texttt{I-TYPE} marks its continuation, and \texttt{O} marks a token
outside any entity. A common variant, BILOU, adds finer tags for the last token of an entity
and for single-token entities. These labels turn the problem of marking spans into a problem
of classifying tokens one by one.

Token labels are not independent, since some label sequences are not valid. An
\texttt{I-DISEASE} label should not follow a \texttt{B-TREATMENT} label, for example, because
the two belong to different entities. Conditional random fields \citep{lafferty2001crf} model
the dependencies between consecutive labels and assign a score to the whole sequence, which
lets them avoid these invalid transitions. They were the standard method for NER before deep
learning. The BiLSTM-CRF model later combined two parts: a bidirectional long short-term
memory network, which builds a contextual representation of each token, and a CRF decoding
layer on top \citep{lample2016ner}. It also used character-level embeddings to capture word
morphology, and it was the state of the art before BERT.

\subsection{BERT-based NER}

With BERT \citep{devlin_bert_2019}, NER becomes a pretrained encoder with a linear
classification layer on each token, and the whole model is fine-tuned end to end. This
approach needs no handcrafted features. It often needs no CRF either, though whether a CRF
layer still helps is debated and the reported gain is usually small. The pretrained encoder
brings general linguistic knowledge to the task through transfer learning, so the model starts
from useful representations rather than from random weights.

The strength of this approach comes from the deep contextual representations that BERT
produces. The same word takes a different vector in a different context, so the word
``cancer'' in ``lung cancer'' and in ``breast cancer'' is represented differently. This
sensitivity to context helps the classifier separate entity types that depend on the
surrounding words. The result is a simple and strong baseline that has become the default for
general-domain NER.

\subsection{Beyond Token-Level Labeling}

Span-based NER takes a different view of the task. Instead of labeling each token, it
enumerates candidate spans of text and classifies each span as an entity of some type or as no
entity. This avoids the constraints of token-by-token tagging and handles nested entities,
where one entity sits inside another. Nested entities are common in biomedicine. In ``right
lung cancer'', an anatomy entity (``right lung'') sits inside a disease entity (``right lung
cancer''), and the BIO scheme cannot give a token two labels at once.

These methods work well on general-domain text, where entities such as people, places, and
organizations follow fairly regular patterns. The biomedical domain raises its own
difficulties, including a large and technical vocabulary, frequent nested entities, negation,
and a telegraphic clinical style, which the next section takes up.
\section{Biomedical and Clinical NER}
\label{sec:ie-biomed-ner}

\subsection{English Benchmarks}

The biomedical NER literature in English rests on a stable set of benchmarks. Scientific-text
NER is mostly evaluated on PubMed abstracts. BC5CDR annotates chemical and disease entities in
1500 PubMed articles \citep{li2016bc5cdr}, NCBI Disease marks disease mentions in 793 abstracts
\citep{dogan2014ncbi}, and JNLPBA tags genes and proteins in the GENIA corpus with five entity
types \citep{collier2004jnlpba}. Other corpora target narrower entity sets, such as AnatEM for
anatomical entities \citep{pyysalo2014anatem} and ChemProt for chemical-protein interactions,
which is a relation-extraction task rather than plain NER \citep{krallinger2017chemprot}. These
datasets cover the language of scientific articles, where sentences are mostly complete and
well edited.

Clinical-text NER is evaluated on a separate family of shared tasks built from hospital notes.
The i2b2 challenges, later renamed n2c2, ran several editions on clinical text. The 2010
i2b2/VA challenge covered concepts, assertions, and relations in clinical records
\citep{uzuner2011i2b2}, and the 2018 n2c2 task addressed adverse drug events and medication
extraction \citep{henry2020n2c2}. A few benchmarks frame biomedical extraction as
classification rather than span labelling, such as GAD for gene-disease associations as a
binary task \citep{bravo2015gad} and the hallmarks-of-cancer multi-label classification
\citep{baker2016hallmarks}. To make these scattered tasks comparable, BLURB collected thirteen
biomedical datasets into one benchmark, and PubMedBERT was the first model evaluated on it
\citep{gu_domain-specific_2022}.

\subsection{French Benchmarks}

French biomedical NER has fewer resources, but several are now standard. QUAERO is the main
reference. It has an EMEA set of drug leaflets and a MEDLINE set of scientific titles, annotated
with ten UMLS semantic groups and including nested entities \citep{neveol14quaero}. CAS
provides clinical cases drawn from scientific articles and was used as task 3 of DEFT 2020,
with a two-class version (CAS1) and an eight-class version (CAS2) \citep{cardon_presentation_2020}.
E3C is a multilingual corpus of clinical cases annotated with a single clinical-entity class
\citep{magnini_e3c_2021}. The CLEF eHealth 2016 campaign combined NER and normalisation on
QUAERO using the BRATeval tool, and added ICD-10 coding of death certificates
\citep{neveol_clinical_2016}.

These corpora were long used in isolation, which made models hard to compare. DrBenchmark
was the first unified French biomedical benchmark, gathering twenty tasks over twelve datasets
so that models could be compared under one protocol \citep{labrak2024drbenchmark}. Its results
show that biomedical pretraining helps on most of these tasks, though no single model is best
on all of them.

\subsection{Specifics of Clinical Text}

Clinical text differs from both scientific text and general web text in ways that affect
extraction. It is telegraphic, with incomplete sentences and lists of symptoms, and it is full
of non-standard abbreviations, such as IDM for myocardial infarction and HTA for arterial
hypertension in French. Negation is central, since a phrase like ``no sign of pneumonia''
mentions pneumonia while ruling it out, and the rule-based NegEx algorithm is still used as a
baseline for detecting it \citep{chapman2001negex}. Temporality also matters, since a model
must separate medical history, current diagnosis, and hypothesis. Expressions of
uncertainty such as ``probable'' or ``to be ruled out'' are common. Spelling errors are
frequent as well, because clinicians type quickly under time pressure.

For these reasons, models trained on scientific or general text perform poorly on clinical
text, which is why domain-pretrained models were developed (discussed in
\Cref{sec:lm-biomedical}). These specialised models help, but they remain supervised, so each
target class still needs annotated examples. One architectural alternative is ClinicalMamba,
which adapts state-space models to clinical text with linear-time complexity in sequence
length, useful for long clinical documents \citep{yang2024clinicalmamba}.

We now have benchmarks and models for biomedical NER. Medical coding is a different task: it
assigns codes from an ontology with thousands of possible labels to a whole document.
\section{Medical Coding}
\label{sec:ie-coding}

\subsection{The Problem}

Medical coding assigns standardised codes to a whole clinical document. The codes come from
controlled systems such as ICD-10 for diagnoses, CCAM for procedures, and ATC for drugs. In
France, the PMSI requires that each hospital stay be coded with a principal diagnosis, a set of
associated diagnoses, and the CCAM procedures performed. This covers about 30 million stays a
year, and the coding is done by hand. The codes feed both hospital funding and national health
statistics, so a coding error has financial and epidemiological consequences.

The task is hard for several reasons. The label space is very large, with about 17{,}000 codes
in the French version of ICD-10, about 8{,}000 CCAM codes, and about 6{,}000 ATC codes. The
documents are long, since a discharge summary runs from 1000 to 3000 tokens, well beyond the
512-token limit of BERT. The code distribution is very skewed, with a few codes that appear
often and a long tail of codes that appear rarely. Finally, a single stay is multi-label, since
one principal diagnosis and several associated diagnoses must be predicted at once.

\subsection{Approaches}

Early work treated coding as multi-label classification with attention over the document.
\citet{mullenbach2018caml} proposed CAML, a convolutional model with one attention head per
label, so each code attends to its own passages of text. This per-label attention is also
explainable, since it shows which words motivated each predicted code.
\citet{cao2020hypercore} extended this idea with HyperCore. It represents the ICD hierarchy
with hyperbolic embeddings, since hyperbolic space fits tree structures better than Euclidean
space, and it adds a graph of code co-occurrences.

Pretrained language models then entered the field. \citet{huang2022plmicd} introduced PLM-ICD,
the first to make pretrained encoders work competitively for ICD coding, with label attention over the model
representations and segment pooling to handle long documents. Its main finding is that a better
pretrained encoder leads to better coding, so pretraining quality matters directly for this
task. \citet{yuan2022msmn} took a different route with MSMN, a multiple-synonym matching
network that uses UMLS synonyms for each code. This helps because clinical text uses varied
expressions for one concept, and a synonym table lets the model match these variants to the
same code.

Large language models do poorly on coding. \citet{soroush2024gpt4coding} benchmarked large
language models for medical coding and found them highly error-prone. Specialised approaches such as PLM-ICD and
MSMN remain stronger, so for coding a specialised encoder still outperforms a generic large
model.

\subsection{Coding in French}

French resources for coding are scarce. \citet{pignat2024fracco} released FRACCO, a French
oncology corpus of 1301 synthetic clinical cases annotated with ICD-O codes. \citet{zaghir2023frasimed}
released FRASIMED, a French adaptation of Spanish coding corpora with about 2000 clinical
documents. The CLEF eHealth 2016 shared task also provided ICD-10 coding of French death
certificates. These corpora are small and cover narrow domains, so they support evaluation
more than large-scale training.

Coding works on English benchmarks such as MIMIC, but the label space is huge, and annotating
examples for every one of the thousands of codes is impossible. French resources are scarcer
still. This motivates approaches that generalise to labels never seen in training, and before
reaching those, we first ask how to connect a mention in the text to a concept.
\section{Entity Linking and Normalization}
\label{sec:ie-linking}

\subsection{The Problem}

Entity linking maps a mention found in text to an entity in a knowledge base such as UMLS,
SNOMED-CT, or ICD-10. The mention is a span of text, and the target is a single concept with a
stable identifier. A related task, normalization, maps the many surface forms of a concept to
one canonical entry. The phrases ``myocardial infarction'', ``MI'', and ``heart attack'' all
describe the same condition, and normalization sends them to a single ICD-10 concept. Both
tasks deal with the gap between how clinicians write and how knowledge bases store information.

Entity linking differs from medical coding in scope. Linking operates on a single mention and
assigns it one concept, whereas coding operates on a whole document and assigns a set of codes
that summarize the encounter. The two tasks
are connected, and linking can be seen as a building block of coding. A typical pipeline first
recognizes entities with named entity recognition, then links each mention to a concept, and
finally aggregates these concepts into document-level codes. Strong linking is therefore
useful well beyond the linking task on its own.

\subsection{Approaches}

Several biomedical encoders are used directly for this task. SapBERT gives strong entity
linking with a single model that covers all entity types \citep{liu2021sapbert}, and CODER
adds relation information to improve term normalization \citep{yuan2022coder}. We described how
both models are pretrained in \Cref{sec:corpus-knowledge}, so here we note only their role as
ready encoders for matching mentions to concepts.

BioSyn takes a more structured view of normalization \citep{sung2020biosyn}. It represents a
mention with both a sparse and a dense encoder and adds their scores to retrieve candidate
concepts. Because one concept often has several synonyms, BioSyn aggregates scores across the
synonyms of a candidate using a marginal objective during training, and it refreshes the hard
candidates as training proceeds. This synonym marginalization is what improves the final
results.

These methods share clear limits. Most of the work is done in English and built on UMLS. UMLS
covers English vocabulary well, but it matches French clinical language poorly, with its
informal abbreviations and local synonyms. The methods also treat each concept in isolation.
They do not use the structure of the ontology, such as parent concepts, exclusions, and
complications, and they ignore the RDF and SKOS relations that connect concepts, including
broader, narrower, related, and exactMatch links.

These methods can normalize a mention once it has been found and once its label exists in
training. But clinical extraction needs to recognize and normalize text across thousands of
concepts never seen during training.
\section{Zero-shot and Open NER}
\label{sec:ie-zeroshot}

\subsection{Why Zero-shot}

Medical label sets are not fixed. New diagnostic codes are added, new clinical trials define
new eligibility criteria, and new diseases appear, as the arrival of COVID-19 showed. A model
that extracts information from clinical text must therefore handle labels that did not exist
when it was trained. Collecting annotated examples for every possible label is not realistic.
The ICD-10 classification alone contains about 17{,}000 codes, and annotating clinical text
for each one would take more effort than any project can afford. Classic named entity
recognition models do not help here, because a BERT model with a linear classification head
has a fixed number of output classes chosen before training. This situation motivates models
that can recognise entity types they have never seen during training.

\subsection{GLiNER}

GLiNER treats named entity recognition as a matching problem between text spans and
natural-language descriptions of entity types \citep{zaratiana2024gliner}. A single
bidirectional transformer encodes the input text and the candidate entity types together. The
types are written as natural language, such as ``disease'' or ``drug name'', rather than as
fixed numeric labels, so a user can ask for any type at inference time. The model places span
representations and type representations in a shared space and matches them by similarity, as
shown in \Cref{fig:ie-gliner}.
GLiNER reaches higher zero-shot scores than ChatGPT while using up to 300 million parameters,
which is far smaller than a large language model. GLiNER2 extends this design into a single
framework that covers named entity recognition, classification, and
structured extraction in one model \citep{zaratiana2025gliner2}. A biomedical version,
GLiNER-BioMed \citep{zaratiana2025glinerbiomed}, fine-tunes the approach on biomedical entity
recognition datasets and reports gains over the general GLiNER on biomedical benchmarks. This
result shows that domain adaptation still helps even for a zero-shot model. The quality of
such models depends heavily on their training data. UniversalNER builds its training set by
asking ChatGPT to annotate text from the Pile, a large general corpus
\citep{zhou2024universalner}. This yields about 45{,}000 instruction examples covering some
13{,}000 entity types, and a model tuned on this data beats ChatGPT on named entity
recognition by several points.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[x=1cm,y=1cm, font=\sffamily,
  sq/.style={draw=ThesisInk!35, fill=white, rounded corners=1.5pt,
    minimum size=0.6cm, inner sep=0pt, text=ThesisInk},
  esq/.style={draw=ThesisTertiaryDark, fill=ThesisTertiary!35, rounded corners=1.5pt,
    minimum size=0.6cm, inner sep=0pt, text=ThesisInk, font=\tiny},
  wlab/.style={rotate=90, font=\scriptsize, text=ThesisInk, inner sep=1pt},
  idx/.style={font=\tiny, text=ThesisNeutral},
  band/.style={draw=ThesisPrimaryDark, fill=ThesisPrimary!16, rounded corners=3pt,
    text=ThesisInk, inner sep=3pt},
  reg/.style={draw=ThesisNeutral!70, densely dashed, rounded corners=4pt, line width=0.6pt},
  conn/.style={ThesisInk!30, line width=0.5pt},
  vcell/.style={draw=ThesisNeutral!55, minimum width=0.42cm, minimum height=0.2cm,
    inner sep=0pt},
  hdr/.style={font=\scriptsize\bfseries, text=ThesisNeutral},
  rlab/.style={font=\scriptsize, text=ThesisNeutral, align=left, anchor=west},
  typechip/.style={draw=ThesisTertiaryDark, fill=ThesisTertiary!30, rounded corners=2pt,
    minimum width=1.5cm, minimum height=0.5cm, inner sep=1pt, text=ThesisInk, font=\scriptsize},
  colhdr/.style={font=\scriptsize, text=ThesisInk, align=center},
  mcell/.style={draw=ThesisInk!25, minimum width=2.55cm, minimum height=0.72cm,
    inner sep=0pt, text=ThesisInk, font=\scriptsize},
  hit/.style={mcell, draw=ThesisSecondaryDark, line width=1pt, fill=ThesisSecondary!42,
    font=\scriptsize\bfseries},
  dasharr/.style={dashed, ThesisPrimaryDark, line width=0.8pt,
    -{Stealth[length=4pt]}, shorten >=2pt, shorten <=2pt},
]
\newcommand{\vbar}[2]{%
  \foreach \sh/\yo in {18/0.3,46/0.1,28/-0.1,60/-0.3}{
    \node[vcell, fill=ThesisNeutral!\sh] at (#1,{#2+\yo}) {};}}
% ============ TOP: region labels + dashed boundaries ============
\node[hdr] at (0.42,2.35) {entity-type prompts};
\node[hdr] at (6.65,2.35) {input sentence};
\draw[reg] (-0.5,-0.95) rectangle (1.35,1.95);
\draw[reg] (1.95,-0.95) rectangle (11.35,1.95);
% ---- prompt units (peach [ENT] square + rotated name) ----
\foreach \i/\w in {0/drug name, 1/disease}{
  \pgfmathsetmacro{\x}{0.85*\i}
  \node[esq] (e\i) at (\x,0) {[ENT]};
  \node[wlab, text=ThesisTertiaryDark] at (\x,1.0) {\w};}
% ---- sentence tokens (neutral square + rotated word + index) ----
\foreach \j/\w in {0/The,1/patient,2/was,3/started,4/on,5/metformin,6/for,7/type,8/2,9/diabetes,10/.}{
  \pgfmathsetmacro{\x}{2.4+0.85*\j}
  \node[sq] (s\j) at (\x,0) {};
  \node[wlab] at (\x,1.0) {\w};
  \node[idx] at (\x,-0.62) {\j};}
% ============ vertical connectors: squares -> transformer ============
\foreach \x in {0,0.85,2.4,3.25,4.1,4.95,5.8,6.65,7.5,8.35,9.2,10.05,10.9}{
  \draw[conn] (\x,-0.32) -- (\x,-1.02);}
% ============ TRANSFORMER band ============
\node[band, minimum width=11.95cm, minimum height=0.72cm] (tf) at (5.425,-1.4)
  {Bidirectional transformer \; (BERT, DeBERTa, \dots)};
% connectors transformer -> token representations
\foreach \x in {0,0.85,2.4,3.25,4.1,4.95,5.8,6.65,7.5,8.35,9.2,10.05,10.9}{
  \draw[conn] (\x,-1.78) -- (\x,-2.42);}
% ============ token / word representations (grayscale vector bars) ============
\foreach \x in {0,0.85,2.4,3.25,4.1,4.95,5.8,6.65,7.5,8.35,9.2,10.05,10.9}{ \vbar{\x}{-2.8} }
\node[rlab] at (11.65,-2.8) {token / word\\ representations};
% ============ FFN + span representation bands ============
% connectors token reps -> bands
\foreach \x in {0,0.85}{ \draw[conn] (\x,-3.22) -- (\x,-4.12); }
\foreach \x in {2.4,3.25,4.1,4.95,5.8,6.65,7.5,8.35,9.2,10.05,10.9}{
  \draw[conn] (\x,-3.22) -- (\x,-4.12); }
\node[band, minimum width=1.9cm, minimum height=0.6cm] (ffn) at (0.42,-4.42) {FFN layer};
\node[band, minimum width=9.4cm, minimum height=0.6cm] (spn) at (6.65,-4.42)
  {span representation layer};
% ============ entity-type embeddings + span embeddings ============
\foreach \x in {0,0.85}{ \draw[conn] (\x,-4.72) -- (\x,-5.22); \vbar{\x}{-5.6} }
\node[rlab, anchor=east, text width=1.7cm, align=right] at (-0.65,-5.6)
  {entity-type\\ embeddings};
\foreach \x in {4.1,6.65,9.2}{ \draw[conn] (\x,-4.72) -- (\x,-5.22); \vbar{\x}{-5.6} }
\node[rlab] at (11.65,-5.6) {span\\ embeddings};
% connectors span embeddings -> matrix column headers
\foreach \x in {4.1,6.65,9.2}{ \draw[conn] (\x,-6.0) -- (\x,-6.55); }
% ============ similarity matrix ============
% column headers
\node[colhdr] at (4.1,-6.82) {metformin\\ {\tiny\itshape\textcolor{ThesisNeutral}{(5,5)}}};
\node[colhdr] at (6.65,-6.82) {type 2 diabetes\\ {\tiny\itshape\textcolor{ThesisNeutral}{(7,9)}}};
\node[colhdr] at (9.2,-6.82) {patient\\ {\tiny\itshape\textcolor{ThesisNeutral}{(1,1)}}};
% row headers (type chips)
\node[typechip] (ra) at (1.7,-7.55) {drug name};
\node[typechip] (rb) at (1.7,-8.30) {disease};
% cells
\node[hit]   at (4.1,-7.55) {0.97};
\node[mcell] at (6.65,-7.55) {0.05};
\node[mcell] at (9.2,-7.55) {0.11};
\node[mcell] at (4.1,-8.30) {0.08};
\node[hit]   at (6.65,-8.30) {0.95};
\node[mcell] at (9.2,-8.30) {0.09};
% dashed connectors: entity-type embeddings -> matrix rows
\draw[dasharr] (0,-6.0) to[out=-90,in=180] (ra.west);
\draw[dasharr] (0.85,-6.05) to[out=-90,in=180] (rb.west);
% matrix caption line
\node[font=\scriptsize, text=ThesisNeutral] at (6.65,-9.0)
  {entity-type / span similarity matrix \; (dot product $+$ sigmoid)};
\end{tikzpicture}}
\caption{GLiNER encodes the entity-type prompts and the input sentence with one
bidirectional transformer, applies a feed-forward layer to the type markers and a span
layer to the token representations, then matches the resulting type and span embeddings in
a similarity matrix whose high-scoring cells are the recognised entities
\citep{zaratiana2024gliner}.}
\label{fig:ie-gliner}
\end{figure}

\subsection{Other Approaches}

GoLLIE takes a generative route. A large language model reads natural-language annotation
guidelines and produces structured annotations that follow them \citep{sainz2024gollie}.
Because the guidelines describe the entity types in words, the model works zero-shot and can
follow detailed definitions, at the cost of being larger and slower than an encoder. NuNER is
a compact named entity recognition foundation model built on a RoBERTa-base encoder
\citep{bogdanov2024nuner}. It is pretrained with contrastive learning on GPT-3.5 annotations of
the C4 web corpus, and its authors find that the size and diversity of this training data
matter more than any other choice. Large language models have also been applied directly to
clinical named entity recognition. Early work was encouraging \citep{agrawal_large_2022},
but \citet{lehman_we_2023} argued that BERT-style models remain
competitive on extractive clinical tasks, that large models are hard to run inside a hospital,
and that sending clinical text to a proprietary API raises confidentiality problems.

\subsection{Limits of Current Zero-shot}

Several open problems remain. GLiNER and similar models are pretrained on general data such as
the Pile, not on clinical text, so they carry the same domain gap that motivated specialised
biomedical models. Entity types are given as short text descriptions, which do not capture the
hierarchy of a medical ontology. The relations between codes, such as parent and child, mutual
exclusion, or synonymy, are not used by the model. Performance also drops on the rare codes
that make up the long tail of medical classifications, where examples are scarcest. Zero-shot
named entity recognition is a promising answer to the problem of thousands of unseen labels.
But the strongest models are trained on general web text, while clinical data are private and
cannot be shared. This raises the question of how to train for the clinical domain without
access to real clinical data.
\section{Synthetic Data for Clinical NLP}
\label{sec:ie-synthetic}

\subsection{The Scarcity Problem}

Clinical text is hard to obtain because it is private. In France, the CNIL regulates the use of
patient data, and in the United States HIPAA plays the same role. A model trained on the
records of one hospital carries information about its patients, so it cannot be shared freely
with another institution. This is the central obstacle for clinical information extraction:
the data that matters most cannot leave the hospital where it was produced. Models and their
training corpora fall under the same rules, which makes collaboration between hospitals
difficult.

Public annotated French clinical resources exist, but they are small and not fully clinical.
Datasets such as QUAERO, CAS, and E3C cover only a few thousand documents, and much of their
text is scientific or drawn from drug leaflets rather than written by clinicians at the
bedside. The larger accessible resources are English. MIMIC, the most widely used clinical
corpus, comes from a single hospital in the United States and reflects its language and coding
practices. For French clinical work there is no large public corpus that matches the style of
real hospital notes, so researchers look for ways to train models without real clinical data.

\subsection{LLM-as-Annotator}

One response uses a large language model to annotate data and then trains a smaller model on
those annotations. The idea appears in several places earlier in this related work. FineWeb-Edu
uses a language model to score documents for educational quality \citep{penedo2024fineweb}.
UniversalNER distils annotations produced by ChatGPT into a smaller model for named entity
recognition \citep{zhou2024universalner}. NuNER distils annotations of web text produced by
GPT-3.5, again for named entity recognition \citep{bogdanov2024nuner}. In each case the large
model supplies labels that would otherwise require human annotators.

For the clinical domain this pattern is attractive for a practical reason. The large model
holds broad medical knowledge and can label text without supervision, while the small model it
teaches is cheap to run and can be deployed inside a hospital where the data must stay. The
expensive model does the labelling once, and the model that is deployed inherits much of its
behaviour. The quality of the small model then depends on how reliable the large model is as an
annotator, which is not guaranteed for technical clinical content.

\subsection{Generating Synthetic Clinical Notes}

A second route generates clinical notes directly with language models, instead of annotating
existing text. \citet{ive2020synthetic} generated synthetic psychiatric records and found
that models trained on the synthetic notes reached performance comparable to models trained on
real notes. This was an early sign that generated text can stand in for real text in clinical
NLP. \citet{chintagunta2021} used GPT-3 to augment medical-dialogue data. A small set of
annotated examples together with generated variants matched the performance of a much larger
annotated set, with an ensemble used to filter low-quality generations. Asclepius pushed the
idea further by training a clinical language model entirely on synthetic notes. It used a large
model to turn public case reports into realistic clinical notes and used no private clinical
data. The result was competitive with models trained on real notes \citep{kweon2024asclepius}.

Synthetic notes built this way still tend to read as too clean and complete, and
they lack the terse, telegraphic style of real clinical writing. The generator can also produce
stereotyped cases that under-represent rare presentations and unusual patients. When a large
model supplies the labels rather than the text, those labels are not gold-standard expert
annotations, and errors in them pass on to the model that learns from them. These caveats
temper the otherwise encouraging finding that generated data can replace real data.

Synthetic data is one answer to the scarcity of clinical text, but it raises a further
question: how do we evaluate a model trained on generated data correctly, and more generally
how do we compare models that use different tokenizers on a fair footing?
\section{Evaluation}
\label{sec:ie-evaluation}

Before any model can be compared with another, we need to agree on how its predictions are
scored. The way we measure performance shapes which model looks best, and the field has not
settled on one way to measure it. This section reviews the metrics used for clinical information
extraction, shows how the choice of tokenizer leaks into the score, and explains why a shared
evaluation protocol is still missing.

\subsection{NER Metrics}

The most common metric for named entity recognition is strict entity-level evaluation,
implemented in the seqeval tool \citep{seqeval}. Under this metric an entity counts as correct
only when its type and its boundaries are both exact, read off from tags in the IOB2 scheme.
Scores can then be aggregated in two ways. The micro-average pools all entities together, so
frequent classes carry the most weight. The macro-average computes a score per class and then
averages those scores, so every class counts equally, however often it appears. Strict
entity-level seqeval has become the de facto standard for named entity recognition, and most
recent French biomedical work reports it.

A second tool, BRATeval, was used in the CLEF eHealth 2016 shared task
\citep{neveol_clinical_2016}. It scores predictions against exact character offsets in the
original text. Because it works on character positions rather than on tokens, it does not
depend on how the text was split or otherwise preprocessed. This makes it robust to
tokenization. It is used far less often than seqeval, so it gives fewer points of
comparison across papers.

A third approach scores each token on its own and reports a weighted F1 over all tokens. Here
every token receives a label and is evaluated separately, instead of being grouped into
entities. The problem is that the ``O'' label, which marks tokens that belong to no entity, is
by far the most frequent label. It therefore dominates the average and inflates the reported score.
This token-level weighted F1 was the metric used by DrBERT \citep{labrak2023drbert}, which
makes its numbers hard to compare with models scored under entity-level seqeval.

\subsection{The Effect of Tokenization}

Token-level metrics carry a further problem. Different tokenizers split the same text into
different numbers of pieces, so a score computed per token depends on the tokenizer used. The
two vocabularies can diverge sharply: SciBERT found only a 42\% overlap between the general
BERT vocabulary and a vocabulary built from scientific text \citep{beltagy_scibert_2019}. A
tokenizer that over-splits technical terms produces more tokens to predict, and therefore more
chances to make an error. The word ``échocardiographie'' is a clear case. A general tokenizer
breaks it into three pieces, while a specialised one breaks it into two. This changes the
number of predictions for the same word, and so changes the score. Token-level metrics are
biased by the choice of tokenizer, and that bias is hard to separate from genuine differences
in model quality.

\subsection{The Need for a Unified Benchmark}

These problems are made worse by the absence of a shared protocol. Each paper tends to use its
own datasets, its own metric, its own train and test splits, and its own hyperparameters. There
is no standard protocol for French biomedical NLP that fixes these choices in advance. Many of
the choices change the result on their own: a CRF layer versus a linear classification head,
frozen versus full fine-tuning of the encoder, and micro versus macro F1 can each move the
ranking of two models. As a result, comparison across published systems is difficult and
often unfair, since differences in the protocol are easily mistaken for differences in the
models.

DrBenchmark was a step toward a shared protocol, gathering several French biomedical tasks under
a common evaluation \citep{labrak2024drbenchmark}. Comparing very different architectures is
harder still. Encoders, large language models, and GLiNER-style bi-encoders use different
tokenizers and different fine-tuning procedures, so a token-level or even an entity-level score
computed on one model's tokenization does not transfer cleanly to another. This points toward
evaluation at the character level, which is independent of the tokenizer and so allows models
of any architecture to be placed on the same scale. Such an evaluation is an open need rather
than a solved problem, and the closing section takes up this missing piece of a fair,
standardised comparison.
\section{Limits and Transition}
\label{sec:ie-limitations}

This chapter reviewed how clinical information extraction is done in French, from the tasks and
corpora to the models that perform them. Across these topics, the same difficulties came back.
These are structural features of the setting, and together they define the questions the rest
of this thesis takes up.

The most immediate difficulty is the scarcity of clinical data. French hospital text is personal data,
and the CNIL places strict conditions on its use. A model trained inside a hospital cannot be
shared with other institutions, so work done on private corpora stays local and hard to
reproduce. The public French resources that remain, such as QUAERO, CAS, and E3C, are small,
and most of their content is scientific writing rather than clinical notes. This shortage of
shareable clinical text pushes the field to look for alternatives to real hospital data.

A second comes from the size of the label space. Clinical coding relies on systems such as
ICD-10 and CCAM, whose tens of thousands of codes (\Cref{sec:ie-coding}) make it infeasible to
annotate examples for every label. A classifier with a fixed output head has one slot per label.
It cannot handle codes it never saw in training, so the classic supervised recipe does not scale
to this setting. The field therefore needs methods that generalise to labels for which no
annotated example exists.

Zero-shot extraction, the apparent way around the label space, is itself not adapted to the
clinic. Models such as
GLiNER and UniversalNER recognise entities described by free-text labels, which removes the
fixed head, but they are trained on general web text. On clinical writing, and especially on
rare codes, their accuracy drops. Closing this gap requires adapting these models to the
biomedical domain.

Ontologies are also underused. Coding systems are published as
structured resources in formats such as RDF and SKOS, with a hierarchy of concepts, exclusion
rules, and lists of synonyms. Zero-shot models reduce each label to a short description and
ignore this structure. The field needs ways to bring ontological knowledge into the
representation of labels.

Finally, evaluation is not standardised. There is no shared benchmark for
French biomedical natural language processing. Token-level scores depend on how each model
splits text, so the tokenizer biases the comparison, and protocols differ from one paper to the
next. Fair comparison calls for an evaluation that does not depend on the tokenizer.

One point cuts across all of these. Much recent attention has gone to large decoder
models, yet for extractive clinical tasks encoder models remain competitive. They are also
small enough to run on the hardware available inside a hospital, which matters when the data
cannot be moved off-site. This keeps encoders at the centre of the work that follows.

These three survey chapters covered the language models used in biomedical natural language
processing, the data used to pretrain them, and the clinical tasks they are built for. The gaps
above set the agenda for the contributions of this thesis. The next part begins with the first
of them, the construction of a biomedical corpus for French from public sources.
