%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Pretraining Data: Corpora and Knowledge}
\label{chap:rw-corpus}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% TODO: le titre du chapitre est provisoire, a discuter


\section{Introduction}
\label{sec:corpus-intro}

Modern language models require very large training corpora. GPT-3 was trained on
about 300 billion tokens~\citep{gpt3}, and the LLaMA models on one to 1.4
trillion~\citep{touvron2023llama}. Scaling-law studies further showed that model
size and training data should grow together for a fixed compute
budget~\citep{chinchilla_scaling}.

Scale alone is not sufficient. The composition and quality of a corpus determine
which language and domains a model learns. Decisions about which documents to
include, remove, or prioritise are therefore part of model design.

French clinical models face a particular shortage of suitable data. French
biomedical text is less abundant than its English counterpart, while hospital
records are private and rarely available for pretraining. This chapter examines
two possible sources of knowledge under these constraints: public text corpora
and structured medical resources such as ontologies.
\section{Web-Scale Pretraining Corpora}
\label{sec:corpus-web}

Large language models are trained on text collected from the open web. Before describing
how this text is cleaned and selected, we describe where it comes from and how the main
public corpora were built.

\subsection{Common Crawl}

Almost every large pretraining corpus starts from Common Crawl, an open crawl of the public
web that has been running since 2008. Common Crawl releases a new snapshot roughly every
month. Each snapshot contains on the order of 2.5 billion web pages and reaches hundreds of
terabytes once uncompressed. The content is very heterogeneous, mixing news articles, forum
threads, spam, source code, advertising, and page boilerplate. For this reason the raw
crawl cannot be used directly for training. Every corpus described below applies a heavy
cleaning pipeline to Common Crawl before the text is suitable for a language model.

\subsection{Early Large Corpora}

The first widely used corpus of this kind was C4, the Colossal Clean Crawled Corpus. It
contains about 750~GB of English text taken from a single Common Crawl snapshot from April
2019, cleaned with simple filters such as language detection, the removal of incomplete
sentences, and deduplication. C4 was introduced together with the T5 model
\citep{raffel2020t5}. \citet{dodge2021documenting} later studied what C4 actually contains,
and found patents, United States government and military sites, machine-generated text, and
even examples drawn from evaluation benchmarks. They also showed that the blocklist used
during cleaning removes a disproportionate amount of text written by and about minorities.

A different design choice is to assemble a corpus from many curated sources rather than from
the web alone. The Pile follows this idea. It contains about 800~GB of English text from 22
sources, including PubMed, arXiv, GitHub, Stack Exchange, Wikipedia, and books. The
motivation is that a diversity of sources matters more for a language model than the raw
volume of any single source. The Pile was open-sourced by EleutherAI \citep{gao2020pile}.

For languages other than English, OSCAR was an important early resource. It is produced by
an efficient pipeline that classifies Common Crawl documents by language and groups them
into per-language corpora. Its French portion, about 138~GB, is the corpus on which
CamemBERT was trained \citep{OrtizSuarezSagotRomary2019}.

\subsection{Modern Corpora}

Recent corpora are much larger and come with carefully documented pipelines. FineWeb
contains about 15 trillion tokens drawn from 96 Common Crawl snapshots. Its pipeline
combines text extraction, language filtering, quality heuristics, and deduplication based
on URLs and MinHash. FineWeb also provides a subset of about 1.3 trillion tokens, called
FineWeb-Edu, that is filtered for educational content, which we discuss in the next section
\citep{penedo2024fineweb}.

Other projects focus on making filtering easier for the people who reuse the corpus.
RedPajama-V2 is an open dataset for training large language models, and it includes a French
subset. It ships pre-computed quality signals, so that users can apply their own filters
without recomputing those signals \citep{weber2024redpajama}. TxT360 instead emphasises
global deduplication. It deduplicates globally across 99 Common Crawl snapshots, leaving roughly 4.83 trillion
tokens \citep{tang2024txt360}.

Two further corpora were released mainly to support open research on language models: Dolma,
about 3 trillion tokens \citep{soldaini2024dolma}, and ROOTS, a 1.6~TB multilingual
composite built for the BLOOM model \citep{laurencon2023roots}.

These corpora give us billions to trillions of tokens of web text, but their quality varies
enormously from one document to the next, which raises the question of how to select the
good documents.
\section{Data Quality and Curation}
\label{sec:corpus-quality}

The previous section showed that web text is abundant but uneven in quality. A monthly
Common Crawl snapshot mixes clean articles with spam, boilerplate, and machine-generated
pages. Training a model on this raw mixture wastes compute on text that teaches little. The
question is therefore how to keep the good documents and remove the rest. Two families of
methods address this question. The first uses simple rules to drop obviously bad text. The
second uses learned classifiers to judge the quality of the content.

\subsection{Heuristic Filtering}

The first family is rule-based filtering, which has been standard practice since the Gopher
pipeline \citep{rae2022gopher}. These rules look at surface properties of a document and
discard it when a property falls outside an expected range. Common rules set minimum and
maximum document lengths, limit the ratio of words to symbols, cap the proportion of
uppercase characters, and remove documents with many duplicated lines. A second rule is
perplexity filtering. A small n-gram model trained on Wikipedia scores each document, and
documents with high perplexity are treated as noise and removed. Language identification
with fastText keeps only documents in the target language. Almost every modern pipeline
reuses these heuristics. They remove the obvious noise, but they do not judge whether the
content is informative.

\subsection{Quality Classifiers}

The second family judges content quality with a learned classifier. The recent pattern has
three steps. A large language model annotates a sample of documents, a small and fast
classifier is distilled from those annotations, and that classifier is then run over the
whole corpus. This keeps the judgement of a strong model while making large-scale filtering
affordable. FineWeb-Edu is a clear example \citep{penedo2024fineweb}. Its authors used
Llama-3-70B-Instruct to score about 460{,}000 web pages for their educational value on a
scale from 0 to 5, distilled a small regression classifier from these scores, and then
classified the full 15 trillion tokens at a cost of about 6{,}000 H100 hours. The filter
removes about 92\% of the data, yet a model trained on the remaining fraction still beats one
trained on the full corpus on benchmarks such as MMLU and ARC.

Other pipelines change the target of the classifier. DataComp-LM, known as DCLM, trains a
fastText classifier to separate instruction-style text from general web text, which favours
pages with a resolved question-and-answer structure \citep{li2024datacomp}. Nemotron-CC uses
a quality classifier with several criteria rather than a single educational score, which
gives a more granular view of quality \citep{nemotron_cc}.

\subsection{Organising by Domain}

Quality is not the only axis along which a corpus can be shaped. WebOrganizer organises the
web into taxonomies by topic and by format, again using language model annotations distilled
into efficient classifiers \citep{wettig2024weborganizer}. This view exposes what quality
scores leave implicit. The authors show that FineWeb-Edu carries a hidden bias toward
science, technology, and academic writing, and that part of its measured gains come from a
domain mix that happens to match the benchmarks. They also find that combining domain
organisation with quality filtering works better than quality filtering alone. What a corpus
contains therefore depends on both how good its documents are and what they are about.

\subsection{Deduplication}

A separate problem is repetition. Web crawls contain the same text many times, because pages
are copied, templated, and re-crawled across snapshots. Training on this material biases the
model toward the most-copied text, such as boilerplate and templates, and wastes capacity on
memorising it. Deduplication addresses this in two stages. The first removes exact duplicates
at the level of URLs. The second finds near-duplicate documents using MinHash together with
locality-sensitive hashing, which compares documents quickly without checking every pair.
The scale of the problem is large. TxT360 deduplicated globally across Common Crawl snapshots
to remove a large volume of near-duplicate documents \citep{tang2024txt360}.

\subsection{Contamination}

Aggressive filtering toward benchmark-like text raises a further concern, benchmark
contamination. Contamination happens when test items, or text that closely resembles them,
leak into the training data. When this occurs, reported scores stop measuring true
generalisation and instead reward memorisation. \citet{deng2024benchmarkprobing} showed that
strong models can guess a large share of masked benchmark options, which is evidence that the
underlying question-and-answer pairs are already on the web. Tools such as Infini-gram help
detect this by finding exact matches between benchmark text and training sets
\citep{liu2024infinigram}, and a survey reviews the broader set of contamination-detection
techniques \citep{xu2024contaminationsurvey}. This leaves a tension in curation.
Filtering more aggressively raises benchmark scores, but some of that gain may come from
implicit contamination rather than from real quality.

These methods work well for general web corpora, where text is abundant and the goal is to
keep the best of a large pool. The biomedical setting is different. The corpora are far
smaller, the domain is specific, and the criteria for good text are not the same, which
raises the prior question of what biomedical sources exist in the first place.
\section{Biomedical and Clinical Corpora}
\label{sec:corpus-biomed}

The data-quality criteria of the previous section were developed for general web text. The
biomedical setting is different, and the corpora used to train biomedical models come from a
small number of recurring sources. We review those sources here, separating the large
English collections, the article-level curation built on top of them, and the more
fragmented French resources. The models trained on these corpora were described earlier in
\Cref{sec:lm-biomedical}; this section is about the corpora themselves.

\subsection{Large English Sources}

The most widely used source is PubMed, the database maintained by the United States National
Library of Medicine. It contains tens of millions of biomedical citations and abstracts,
written in technical vocabulary and edited to a consistent standard. Its main limit is that
it provides abstracts only, without the full text of the articles. PubMed Central (PMC) Open
Access addresses this by providing full text for millions of articles
\citep{chen2023meditron}. The biomedical text drawn from it is overwhelmingly English, about
98\% in one large sample \citep{labrak-etal-2024-biomistral}. The collection is heterogeneous,
mixing research articles, reviews, case reports, editorials, and clinical guidelines. It is
also hard to use at the article level, because a single research article can combine
clinically relevant case material with methodological sections that are of little clinical use.

Clinical text is harder to obtain. The main reference is MIMIC
\citep{johnson2016mimiciii,johnson_mimiciv_2023}, a set of clinical notes from a single
United States hospital, the Beth Israel Deaconess Medical Center. Access is restricted and
requires a data use agreement and ethics training. MIMIC is treated as a gold standard for
English clinical text, but it comes from one hospital in one country, which limits how far
models trained on it generalise to other settings.

\subsection{Article-Level Curation for Biomedicine}

Some work tries to improve these corpora by scoring articles before training. The Meditron
work \citep{chen2023meditron} evaluates an article-level scheme that scores articles by
features such as their MeSH subject tags, publication type, journal, recency, and citation
count, upsampling high-scoring articles and filtering out low-scoring ones. The same limit
returns here, because the scoring acts on whole articles and so still mixes relevant and
non-relevant content within a single article.

A second example is the corpus used for BioClinical ModernBERT
\citep{sounack2025bioclinicalmodernbert}, which combines PubMed, PMC, and around twenty
clinical datasets into one large collection. Most of those clinical sources require their own
data use agreements, so the assembled corpus is not fully public. The model
built on it is discussed in \Cref{sec:lm-biomedical}; what matters here is that the access
constraint sits in the clinical part of the corpus.

\subsection{French Sources}

Public French biomedical resources exist, but they are scattered across several projects.
ISTEX is a base of tens of millions of scientific references from which French biomedical
documents can be extracted. CLEAR provides French drug leaflets and medical encyclopaedia
articles in both technical and simplified versions \citep{grabar_clear_2018}. E3C is a
multilingual corpus of clinical cases, drug leaflets, and PubMed clinical-case abstracts
\citep{magnini_e3c_2021}. HAL, the French open repository, and its harvested version HALvest
provide French and English scientific articles \citep{kulumba2024halvest}. Biomedical
Wikipedia adds a free and multilingual source. Together these resources are useful, but
each is partial, and none approaches the scale of PubMed or PMC.

The French biomedical models, including the ones trained from scratch and the ones obtained
by continual pretraining, were described in \Cref{sec:lm-biomedical}; here we note only the
corpora they used and the pattern of results. The continual-pretraining attempts on French
biomedical text relied on small or loosely targeted corpora. One continued pretraining on about 31{,}000 abstracts
and reported no improvement with the large model \citep{copara_contextualized_2020}. Another assembled about 136
million words from PubMed, Cochrane, ISTEX, and Wikipedia, and obtained a small gain on drug
leaflets and no gain on scientific titles \citep{le_clercq_de_lannoy_strategies_2022}. A
third used 21 million private hospital documents, which improved a private clinical benchmark
but could not be shared with other researchers \citep{dura_learning_2022}.

These French results are mixed, and their cause is not settled. We do not know whether the
disappointing outcomes come from the corpora, which may be too small or not well targeted, or
from the continual-pretraining method. These difficulties sharpen the clinical scarcity
introduced earlier. Clinical text is private, and in France it is governed by the data-protection
rules of the CNIL. Models trained inside one hospital cannot be shared freely between institutions.
And the public French sources are fragmented and small next to PubMed and PMC.

Biomedical corpora help models learn the vocabulary and style of medical text. But language
models capture statistical patterns and not the explicit structure that tasks such as medical
coding require, so we now turn to the medical ontologies that provide it.
\section{Medical Ontologies and Knowledge Graphs}
\label{sec:corpus-ontologies}

\subsection{Main Medical Ontologies}

An ontology is a structured vocabulary. It records concepts, the names used to refer to
them, and the relations between them, such as hierarchy, synonymy, and association. It
therefore makes explicit the knowledge that a language model would otherwise have to
learn on its own from running text. A diagnosis, the procedures that treat it, and the drugs
prescribed for it are recorded as concepts and links, rather than left to be inferred from
word statistics. Several large ontologies of this kind cover the biomedical domain, and a few
of them are central to clinical work in France.

The Unified Medical Language System (UMLS) is a large collection of biomedical vocabularies
maintained by the United States National Library of Medicine (NLM) \citep{lindberg_unified_1993}.
It brings together roughly two hundred source vocabularies,
including SNOMED-CT, the Medical Subject Headings (MeSH), the International Classification of
Diseases (ICD), and the Anatomical Therapeutic Chemical classification (ATC). Recent releases contain
several million concepts and over ten million distinct concept names. It records synonyms,
hierarchical and associative relations, and definitions, with coverage that varies by source.

SNOMED-CT holds hundreds of thousands of clinical concepts. They are connected by hierarchical
relations, of the is-a kind, and by associative relations such as finding site and causative
agent. It is an international standard for clinical information systems and covers diagnoses,
procedures, anatomy, substances, and organisms.

The International Classification of Diseases, tenth revision (ICD-10, in French CIM-10), is
the World Health Organization's (WHO) disease classification. It is organised as a hierarchy of
chapters, blocks, categories, and subcategories. The French version, maintained by the Agence
Technique de l'Information sur l'Hospitalisation (ATIH), adds inclusions, exclusions, and
notes, and contains tens of thousands of codes. It is used for hospital coding under the French
case-mix system, the Programme de Médicalisation des Systèmes d'Information (PMSI).

The Classification Commune des Actes Médicaux (CCAM) is the French common classification of
medical procedures. It is specific to France and arranges about 8{,}000 codes in a hierarchy.

The Anatomical Therapeutic Chemical classification (ATC) is the WHO classification of drugs
over five levels, from the anatomical group down to the individual substance. It contains
about 6{,}000 codes at the fifth level.

\subsection{Formal Representations}

Many ontologies are represented in a small set of standard formats defined by the World Wide
Web Consortium (W3C), though some clinical terminologies use resource-specific formats. The Resource Description Framework (RDF) stores facts as
subject-predicate-object triples. The Web Ontology Language (OWL) adds description logic,
which allows automatic reasoning over classes and properties. The Simple Knowledge
Organization System (SKOS) provides a lighter vocabulary for thesauri. It uses prefLabel and
altLabel for a concept's preferred and alternative names, and broader, narrower, and related
links for hierarchy and association.

These formats let an ontology state explicitly what a language model must otherwise infer
from text. They record hierarchy, so that type 2 diabetes is marked as a kind of diabetes
mellitus. They record exclusions, so that type 2 diabetes is kept distinct from type 1 and is
not confused with it. They record synonymy, so that one concept carries several labels at
once. They also record associative relations, such as the complications of a disease, its
treatments, and links that cross from one ontology to another.

This structured knowledge is rich, but it is stored as graphs rather than as the running text
that language models expect, which raises the question of how to bring it into pretraining.
\section{Knowledge-Enhanced Pretraining}
\label{sec:corpus-knowledge}

Several lines of work try to close this gap. They differ in how much of the graph they bring
into the model and at what stage of training they do it.

\subsection{Static Graph Embeddings}

The first family turns a graph into fixed vectors, one per node, without textual context. Node2Vec
learns node representations from biased random walks and a skip-gram objective
\citep{grover2016node2vec}. RDF2Vec instead derives sequences from traversals of directed RDF
graphs and may retain relation predicates in those sequences before applying a word2vec-style
objective \citep{ristoski2016rdf2vec}. It became the basis for later path-based ontology
methods. Snomed2Vec
evaluates graph-embedding methods on SNOMED-CT and reports large gains on concept similarity
\citep{agarwal2019snomed2vec}. OWL2Vec* extends the approach by running walks over OWL
ontologies \citep{chen2021owl2vec}. These methods share one limit. The embeddings are static
and live outside any transformer, so a concept gets a single fixed vector whatever the
surrounding text says. \Cref{fig:corpus-static} shows this pipeline.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[font=\scriptsize\sffamily,
  gnode/.style={draw=ThesisInk!35, rounded corners=2pt, fill=white, inner sep=3pt,
    align=center, text=ThesisInk, minimum width=2.1cm, minimum height=0.72cm},
  elab/.style={font=\scriptsize\itshape, text=ThesisNeutral},
  gedge/.style={-{Stealth[length=3pt]}, ThesisNeutral!70, line width=0.7pt, shorten >=1.5pt, shorten <=1.5pt},
  xedge/.style={-{Stealth[length=3pt]}, ThesisPrimaryDark!75, dashed, line width=0.8pt, shorten >=1.5pt, shorten <=1.5pt},
  walkchip/.style={draw=ThesisPrimaryDark, fill=ThesisPrimary!16, rounded corners=2pt,
    text=ThesisInk, align=center, inner sep=2.5pt, minimum height=0.62cm, minimum width=1.7cm},
  walk/.style={-{Stealth[length=3.5pt]}, ThesisPrimaryDark, line width=1.1pt, shorten >=1.5pt, shorten <=1.5pt},
  flow/.style={-{Stealth[length=4pt]}, ThesisInk!40, line width=0.9pt, shorten >=3pt, shorten <=3pt},
  model/.style={rounded corners=3pt, draw=ThesisPrimaryDark, fill=ThesisPrimary!16,
    text=ThesisInk, align=center, inner sep=5pt, minimum height=1.1cm},
  vec/.style={rounded corners=2pt, draw=ThesisSecondaryDark, fill=ThesisSecondary!24,
    text=ThesisInk, minimum width=1.9cm, minimum height=0.42cm, inner sep=2pt, align=center},
  hdr/.style={font=\scriptsize\sffamily\bfseries, text=ThesisNeutral},
]
\newcommand{\code}[1]{{\ttfamily\tiny\textcolor{ThesisNeutral}{#1}}}
% --- Stage 1: illustrative cross-ontology graph ---
\node[gnode] (dm)  at (0,3.9) {Diabetes mellitus\\ \code{ICD-10 E10--E14}};
\node[gnode] (t2)  at (0,2.6) {Type 2 diabetes\\ mellitus\ \code{ICD-10 E11}};
\node[gnode] (met) at (0,1.3) {Metformin\\ \code{ATC A10BA02}};
\node[gnode] (big) at (0,0.0) {Biguanides\\ \code{ATC A10BA}};
\draw[gedge] (t2) -- (dm) node[elab, midway, right=2pt] {is-a};
\draw[xedge] (met) -- (t2) node[elab, midway, right=2pt, text=ThesisPrimaryDark] {treats};
\draw[gedge] (met) -- (big) node[elab, midway, right=2pt] {is-a};
\node[hdr, align=center] at (0,4.7) {illustrative cross-\\ontology graph};
\node[elab, align=center] at (0,-0.72) {dashed: illustrative\\clinical association};
% --- Stage 2: sampled graph path ---
\def\wx{3.4}
\node[walkchip] (w1) at (\wx,3.0) {Metformin};
\node[walkchip] (w2) at (\wx,1.65) {Type 2 diabetes\\ mellitus};
\node[walkchip] (w3) at (\wx,0.3) {Diabetes\\ mellitus};
\draw[walk] (w1) -- (w2) node[elab, midway, right=2pt, text=ThesisPrimaryDark] {treats};
\draw[walk] (w2) -- (w3) node[elab, midway, right=2pt] {is-a};
\node[hdr, align=center] at (\wx,4.7) {sampled\\ graph path};
\node[elab, align=center] at (\wx,-0.72) {relations may also be retained\\in RDF2Vec sequences};
\draw[flow] (met.east) to[out=0,in=180] (w1.west);
% --- Stage 3: skip-gram ---
\node[model] (sg) at (6.3,1.5) {word2vec-style\\ objective};
\draw[flow] (w2.east) -- (sg.west);
\draw[flow] (w3.east) -- (sg.west);
% --- Stage 4: fixed vectors ---
\def\vx{9.0}
\node[vec] (v1) at (\vx,3.0) {Diabetes mellitus};
\node[vec] (v2) at (\vx,2.0) {Type 2 diabetes m.};
\node[vec] (v3) at (\vx,1.0) {Metformin};
\node[vec] (v4) at (\vx,0.0) {Biguanides};
\draw[flow] (sg.east) -- (v2.west);
\draw[flow] (sg.east) -- (v3.west);
\node[hdr, align=center] at (\vx,4.7) {one fixed vector\\ per node};
\end{tikzpicture}}
\caption{Static graph-embedding methods derive sequences from graph traversals and learn one
fixed vector per node with a word2vec-style objective, using biased random walks for Node2Vec
and directed RDF traversals that may retain relation predicates for RDF2Vec
\citep{grover2016node2vec,ristoski2016rdf2vec}.}
\label{fig:corpus-static}
\end{figure}

\subsection{Injecting Knowledge into Transformers}

A second family couples textual representations with explicit entities or graph structure. ERNIE
aligns entity embeddings with text representations during pretraining \citep{zhang2019ernie}.
KnowBERT inserts entity linkers as extra attention layers within BERT
\citep{peters2019knowbert}. K-BERT injects knowledge-graph
triples into the input using soft-position information, so that the added triples do not
disturb the sentence order \citep{liu2020kbert}. A later generation pretrains on language and
knowledge together. KEPLER trains masked language modeling and a knowledge-embedding objective
at the same time on Wikidata, with one shared encoder \citep{wang2021kepler}. CoLAKE builds
word-knowledge graphs that unify text and knowledge context, then pretrains a modified masked
objective on them \citep{sun2020colake}. DRAGON instead encodes a text segment and a relevant
knowledge subgraph as separate modalities, fuses them bidirectionally at multiple layers, and
pretrains with masked language modeling and link prediction \citep{yasunaga2022dragon}.
\Cref{fig:corpus-inject} illustrates K-BERT's input-level mechanism; DRAGON does not use this
layout.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[font=\scriptsize\sffamily,
  tok/.style={draw=ThesisInk!30, rounded corners=2pt, fill=white, text=ThesisInk,
    inner sep=3pt, minimum height=0.6cm},
  ent/.style={tok, draw=ThesisPrimaryDark, fill=ThesisPrimary!22},
  gnode/.style={draw=ThesisInk!35, rounded corners=2pt, fill=white, inner sep=2.5pt,
    align=center, text=ThesisInk, minimum height=0.62cm},
  elab/.style={font=\scriptsize\itshape, text=ThesisNeutral},
  gedge/.style={-{Stealth[length=3pt]}, ThesisNeutral!75, line width=0.7pt, shorten >=1.5pt, shorten <=1.5pt},
  inject/.style={-{Stealth[length=4pt]}, ThesisPrimaryDark, line width=1.1pt, shorten >=2pt, shorten <=2pt},
  flow/.style={-{Stealth[length=4pt]}, ThesisInk!40, line width=0.9pt, shorten >=3pt, shorten <=3pt},
  model/.style={rounded corners=4pt, draw=ThesisPrimaryDark, fill=ThesisPrimary!14,
    text=ThesisInk, align=center, minimum height=1.0cm},
  prep/.style={rounded corners=3pt, draw=ThesisSecondaryDark, fill=ThesisSecondary!18,
    text=ThesisInk, align=center, minimum height=0.86cm},
]
\newcommand{\code}[1]{{\ttfamily\tiny\textcolor{ThesisNeutral}{#1}}}
% transformer band
\node[model, minimum width=10.4cm] (tf) at (4.7,3.35) {Transformer\\ \footnotesize contextual representations};
\node[prep, minimum width=10.4cm] (prep) at (4.7,1.75) {flattened sequence\\
  \footnotesize soft positions + visibility matrix};
% token row
\def\tx{1.35}
\node[tok] (k0) at (0*\tx,0) {The};
\node[tok] (k1) at (1*\tx,0) {patient};
\node[tok] (k2) at (2*\tx,0) {was};
\node[tok] (k3) at (3*\tx,0) {started};
\node[tok] (k4) at (4*\tx,0) {on};
\node[ent] (k5) at (5*\tx,0) {metformin};
\node[tok] (k6) at (6*\tx,0) {.};
\foreach \k in {k0,k1,k2,k3,k4,k5,k6} \draw[ThesisInk!25, line width=0.7pt] (\k.north) -- (\k.north |- prep.south);
\draw[flow] (prep.north) -- (tf.south);
% K-BERT knowledge branch
\node[gnode, draw=ThesisPrimaryDark, fill=ThesisPrimary!12] (rel) at (7.7,-1.45) {treats};
\node[gnode, draw=ThesisPrimaryDark, fill=ThesisPrimary!12] (t2) at (9.8,-1.45) {Type 2 diabetes\\ mellitus\ \code{ICD-10 E11}};
\draw[inject] (k5.south) to[out=270,in=150] (rel.west);
\draw[gedge] (rel) -- (t2);
\draw[flow] (t2.north) to[out=90,in=0] (prep.east);
\node[elab, text=ThesisPrimaryDark, align=center] at (8.45,-2.25) {knowledge branch attached\\to the entity mention};
\end{tikzpicture}}
\caption{K-BERT attaches a knowledge-graph branch to an entity mention and flattens the
resulting sentence tree with soft positions and a visibility matrix before the transformer
encodes it, while other knowledge-enhanced models use different integration mechanisms
\citep{liu2020kbert}.}
\label{fig:corpus-inject}
\end{figure}

\subsection{Medical Contrastive Approaches}

A third family uses contrastive learning to pull related medical terms together in the
embedding space. SapBERT does self-alignment pretraining on UMLS synonyms. It treats synonyms
as positive pairs and terms from different UMLS concepts as negatives, scales to millions of
concepts, and reaches strong biomedical entity-linking results with a single model for all entity types
\citep{liu2021sapbert}. CODER adds contrastive supervision from term--relation--term triples,
injecting relational knowledge beyond synonym alignment \citep{yuan2022coder}.
\Cref{fig:corpus-contrastive} gives a schematic two-dimensional projection of SapBERT's
synonym-alignment objective; placements and distances are illustrative, and CODER's relation-level
objective is not shown.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\scriptsize\sffamily,
  term/.style={draw=ThesisInk!30, rounded corners=2pt, fill=white, text=ThesisInk,
    inner sep=3pt, align=center},
  synonym/.style={term, draw=ThesisSecondaryDark, fill=white},
  neg/.style={term, draw=ThesisTertiaryDark, fill=ThesisTertiary!24},
  lbl/.style={font=\scriptsize\sffamily\bfseries},
  note/.style={font=\scriptsize\itshape, text=ThesisNeutral},
]
% Schematic plane and same-concept neighbourhood.
\begin{scope}[on background layer]
  \fill[ThesisInk!3, rounded corners=10pt] (-4.8,-1.8) rectangle (4.8,1.8);
  \draw[ThesisInk!18, rounded corners=10pt, line width=0.7pt]
    (-4.8,-1.8) rectangle (4.8,1.8);
  \fill[ThesisSecondary!14] (-2.55,0.0) ellipse [x radius=1.95cm, y radius=1.25cm];
\end{scope}
\node[synonym] at (-3.45,0.5) {T2DM};
\node[synonym] at (-1.7,0.55) {NIDDM};
\node[synonym] at (-2.55,-0.45) {diab\`ete de type 2};
\node[lbl, text=ThesisSecondaryDark] at (-2.55,-1.35) {same UMLS concept};
\node[neg] at (2.75,0.15) {Type 1 diabetes\\ mellitus};
\node[lbl, text=ThesisTertiaryDark, align=center] at (2.75,-0.95)
  {different but related concept\\ negative pair};
\node[lbl, text=ThesisNeutral] at (0,2.35) {illustrative 2D embedding projection};
\node[note, anchor=north east] at (4.55,1.55) {not to scale};
\end{tikzpicture}
\caption{Schematic two-dimensional projection of SapBERT's objective, where synonymous terms
associated with the same UMLS concept occupy a compact neighbourhood while a term from a
different concept is separated, with placements and distances illustrative and not to scale
\citep{liu2021sapbert}.}
\label{fig:corpus-contrastive}
\end{figure}

\subsection{Synthetic Data for Pretraining}

A recent direction uses synthetic data to make pretraining or continued pretraining more
data-efficient. The ``Textbooks Are All You Need'' work showed that high-quality,
textbook-like synthetic data can support strong small models, suggesting that data structure
and quality matter alongside raw volume \citep{gunasekar2023textbooks}. EntiGraph addresses
continued pretraining from a small source corpus. It extracts salient entities from each
source document, samples groups of those entities, and prompts a language model with access
to the source document to generate text analysing their relations. In its main experiment,
about 1.3 million real tokens produced about 455 million synthetic tokens, with accuracy
scaling roughly log-linearly in the number of synthetic tokens. The method targets decoder
models and has not yet been tested on bidirectional encoders \citep{yang2025entigraph}.
\Cref{fig:corpus-g2t} shows this route.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[font=\scriptsize\sffamily,
  gnode/.style={draw=ThesisInk!35, rounded corners=2pt, fill=white, inner sep=3pt,
    align=center, text=ThesisInk, minimum height=0.66cm, minimum width=2.2cm},
  elab/.style={font=\scriptsize\itshape, text=ThesisNeutral},
  flow/.style={-{Stealth[length=4pt]}, ThesisInk!40, line width=0.9pt, shorten >=3pt, shorten <=3pt},
  model/.style={rounded corners=4pt, draw=ThesisPrimaryDark, fill=ThesisPrimary!16,
    text=ThesisInk, align=center, inner sep=5pt, minimum height=1.0cm},
  source/.style={draw=ThesisInk!35, fill=white, rounded corners=3pt,
    text=ThesisInk, align=left, inner sep=7pt, text width=3.1cm},
  gen/.style={draw=ThesisSecondaryDark, fill=ThesisSecondary!18, rounded corners=3pt,
    text=ThesisInk, align=left, inner sep=7pt, text width=3.8cm},
  hdr/.style={font=\scriptsize\sffamily\bfseries, text=ThesisNeutral},
]
\newcommand{\code}[1]{{\ttfamily\tiny\textcolor{ThesisNeutral}{#1}}}
% Four equally spaced columns, sharing one main-flow baseline.
\def\colsrc{0}
\def\colent{4.2}
\def\colrel{8.4}
\def\coltxt{12.6}
\def\flowy{0}
\def\heady{2.5}
% source document
\node[source] (src) at (\colsrc,\flowy) {\itshape The source document mentions type 2 diabetes and
metformin in context.};
\node[hdr] at (\colsrc,\heady) {small source document};
% entity extraction and sampling
\node[model] (extract) at (\colent,1.25) {entity\\ extraction};
\node[gnode] (egroup) at (\colent,\flowy) {Type 2 diabetes\\ + Metformin};
\draw[flow] (src.east) to[out=10,in=180] (extract.west);
\draw[flow] (extract.south) -- (egroup.north);
\node[hdr, align=center] at (\colent,\heady) {salient entities};
\node[elab, align=center] at (\colent,-0.95) {sampled pair or triple};
% relation analysis conditioned on both entities and source
\node[model, minimum width=3.4cm, minimum height=1.7cm] (lm) at (\colrel,\flowy)
  {LLM relation analysis\\ \footnotesize conditioned on the\\ \footnotesize source document};
\draw[flow] (egroup.east) -- (lm.west);
\draw[flow] (src.south) to[out=270,in=270,looseness=0.75] (lm.south);
\node[hdr] at (\colrel,\heady) {relation analysis};
% generated text
\node[gen] (txt) at (\coltxt,\flowy) {\itshape The document discusses metformin as a treatment
in relation to type 2 diabetes.};
\draw[flow] (lm.east) -- (txt.west);
\node[hdr, text=ThesisSecondaryDark] at (\coltxt,\heady) {relation-focused synthetic text};
\node[elab, align=center] at (\coltxt,-1.2) {\code{1.3M} real tokens $\rightarrow$ \code{455M} synthetic tokens};
\end{tikzpicture}}
\caption{EntiGraph extracts salient entities from a small source corpus, samples groups of
entities, and prompts a language model conditioned on the source document to generate
relation-focused synthetic text, turning about 1.3 million source tokens into about 455
million synthetic tokens in its main experiment \citep{yang2025entigraph}.}
\label{fig:corpus-g2t}
\end{figure}

Several gaps remain open across these families. Most of the work is in English and built on
UMLS, so other languages and other knowledge sources remain comparatively underexplored.
Entity-centric synthetic augmentation for continued pretraining is recent, with the main
results appearing between 2023 and 2025, and it has so far targeted decoder models rather
than bidirectional encoders. National ontologies, such as the French ICD-10 maintained by
the ATIH and the CCAM procedure coding, have seen little use in pretraining.
Contrastive methods such as SapBERT and CODER use UMLS through term pairs and relation triples
rather than by synthesising relation-focused text from source documents. These open problems
set up the limits we discuss next.
\section{Limits}
\label{sec:corpus-limitations}

The first problem is scarcity outside English. The biomedical text drawn from PMC is about 98\% English
\citep{labrak-etal-2024-biomistral}, and French biomedical corpora remain small and
scattered. The main filtering pipelines that made web-scale curation work, such as
FineWeb-Edu and DCLM, were built for English \citep{penedo2024fineweb,li2024datacomp}. How
to collect and filter text for a specific domain in a language with little data is still an
open question.

The second problem is the granularity of filtering. Existing biomedical curation works at
the document or article level, as in Meditron \citep{chen2023meditron}, which leaves the
mix of relevant and irrelevant content within an article untouched. Finer, paragraph-level
filtering has received little attention.

The third problem is that quality criteria do not transfer across domains. The educational
criterion used by FineWeb-Edu does not match biomedical text such as clinical cases, drug
leaflets, and protocols. The domain biases identified by WebOrganizer
\citep{wettig2024weborganizer} may also take a different shape in the biomedical case. The
right domain-specific quality criteria for biomedical text are still unknown.

Two further gaps concern knowledge. Research on medical ontologies is dominated by UMLS, but
French hospital systems use ICD-10 as maintained by the ATIH, along with CCAM for procedures
and ATC for drugs. We did not find prior pretraining work that uses these French national
coding resources, which matters for medical coding in French. Generating textbook-like text
from such resources to continue pretraining a French encoder, and evaluating it on the target
clinical tasks, is the specific combination we take up later.
