%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Language Models}
\label{chap:rw-lm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:lm-intro}

Long before there were computers, there was the idea that the next symbol in a
text can be guessed from the ones before it. In 1913 the mathematician Andrei
Markov tested it by hand. He took twenty thousand letters from Pushkin's
\emph{Eugene Onegin}, reduced each to a vowel or a consonant, and counted: a vowel
followed another vowel only 13\% of the time, but followed a consonant 66\% of the
time \citep{markov_onegin_1913}. Letters are not independent. What comes before
shapes what comes next, and that can be measured.

Claude Shannon turned this counting into a theory. He modelled a text as a random
source and defined its \emph{entropy}, the average information carried by each
symbol, as
\[
  H = -\sum_i p_i \log_2 p_i,
\]
measured in bits \citep{shannon_mi}. The quantity is largest when every symbol is
equally likely and shrinks as some symbols become predictable.

Shannon showed this with a simple demonstration. Generate letters independently and
the result is noise, \emph{xfoml rxkhrjffjuj zlpwcfwkcyj}. Keep three-letter
statistics and the output is already pronounceable and nearly English, \emph{in no
ist lat whey cratict froure birs}. Stitch whole words together by pairs and the
fragments start to read like real sentences, \emph{the head and in frontal attack
on an english writer that the character of this point}. Each step uses a little
more context, and each step looks more like the language.

How predictable is English, then? Shannon answered with an experiment in which a
person guessed the next letter of a text one symbol at a time
\citep{shannon_prediction_1951}. Shown the preceding text, his reader guessed
correctly about two times in three, usually on the first try; from how the guesses
fell, Shannon derived bounds on the entropy. An alphabet of 27
symbols, the 26 letters and a space, could in principle carry $\log_2 27 \approx
4.76$ bits per character. Yet
given enough preceding text, Shannon estimated the true entropy of English at
about one bit per character, somewhere between roughly 0.6 and 1.3
(\Cref{fig:lm-entropy}). Nearly four of every five possible bits are redundant,
already fixed by what came before. That redundancy is the whole opportunity: a
model that predicts the next symbol well has learned how the language works.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.74\columnwidth, height=4.8cm,
  ybar, bar width=15pt,
  ymin=0, ymax=5.3,
  ylabel={bits per character},
  ylabel style={font=\small},
  symbolic x coords={0,1,2,{$\approx$100}},
  xtick=data,
  xlabel={letters of preceding context},
  xlabel style={font=\small},
  tick label style={font=\small},
  axis lines=left,
  enlarge x limits=0.20,
  every axis plot/.style={fill=ThesisPrimary, draw=ThesisPrimaryDark, line width=0.6pt},
]
\addplot+[error bars/.cd, y dir=both, y explicit]
  coordinates {(0,4.76) +- (0,0) (1,4.14) +- (0,0) (2,3.56) +- (0,0) ({$\approx$100},0.95) +- (0,0.35)};
\end{axis}
\end{tikzpicture}
\caption{The entropy of printed English falls as more preceding context is used,
from $\log_2 27 \approx 4.76$ bits per character with no context to about one bit
(between 0.6 and 1.3) once roughly a hundred letters are taken into account
\citep{shannon_prediction_1951}.}
\label{fig:lm-entropy}
\end{figure}

This is what a language model is. By the chain rule, the probability of a whole
text factorises exactly into a product of next-symbol predictions,
\[
  P(w_1,\dots,w_T) = \prod_{t=1}^{T} P(w_t \mid w_{<t}),
\]
so modeling language reduces to estimating a single conditional distribution: the
probability of the next token given those before it (\Cref{fig:lm-nexttoken}).
Every model in this chapter is an estimator of that one quantity, and the history
of the field is the history of estimating it better.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  ctx/.style={draw=ThesisPrimaryDark, fill=ThesisPrimary!18, rounded corners=2pt,
    minimum width=0.95cm, minimum height=0.6cm, font=\small\sffamily, text=ThesisInk},
  pred/.style={draw=ThesisSecondaryDark, fill=ThesisSecondary!28, rounded corners=2pt,
    minimum width=0.95cm, minimum height=0.6cm, font=\small\sffamily, text=ThesisInk},
]
\node[ctx] (w1) at (0,0) {The};
\node[ctx] (w2) at (1.1,0) {cat};
\node[ctx] (w3) at (2.2,0) {sat};
\node[pred] (w4) at (3.55,0) {?};
\draw[->, ThesisInk!45, shorten >=2pt, shorten <=2pt] (w3.east) -- (w4.west);
\node[font=\scriptsize\sffamily, text=ThesisNeutral] at (1.1,-0.72) {context $w_{<t}$};
\node[font=\scriptsize\sffamily, text=ThesisNeutral] at (3.55,-0.72) {next token $w_t$};
\end{tikzpicture}
\caption{A language model predicts each token from the ones before it. The
probability of a whole text is the product of these next-token predictions.}
\label{fig:lm-nexttoken}
\end{figure}

Estimating that conditional is harder than it sounds. The simplest idea, the one
Markov and the early n-gram models used, is to count: see how often each short
sequence is followed by each next word. But counting runs out almost at once. Most
word sequences never appear twice, even in a large corpus, so the counts are zero
exactly where we need them. Much of the history of language modeling is the search
for estimators that get around this problem, and the rest of this chapter follows
that search, from n-gram counts to neural networks to the Transformer.

For this thesis the point is practical. Clinical information extraction rests
entirely on these models: they are pretrained on large corpora and then adapted to
a task. Adaptation is the hard part, because the target combines three scarcities
at once, French, biomedical, and clinical, and, as the previous part showed, the
text we can gather most easily is the least like the text we need. This chapter
reviews the methods that make this adaptation possible: the architectures and
pretraining objectives, how they scale, how they are specialised to a domain
through continual pretraining, and the tokenization and long-context choices that
matter for clinical documents. Continual pretraining, the mechanism at the heart of
this thesis, is treated last and in the most detail.


\section{From Statistical to Neural Language Models}
\label{sec:lm-stat-neural}

\subsection{N-gram models}

The first estimators keep only a short history. Markov's assumption is that the next
token depends only on the last few, so the full conditional $P(w_t \mid w_{<t})$ is
approximated by $P(w_t \mid w_{t-n+1}, \ldots, w_{t-1})$, the probability of a token
given the $n-1$ before it. These probabilities are read straight from counts in a
training corpus. The trouble is the one raised above, only sharper: the number of
possible $n$-grams grows as $|\mathcal{V}|^{n}$, so for any real vocabulary most
$n$-grams are never seen and their counts are zero. Smoothing methods, of which
Kneser-Ney smoothing \citep{kneser_ney} is the best known, move some probability
onto unseen sequences and fall back to shorter contexts when a longer one is
missing. They made n-gram models good enough to be the state of the art for years.
But they cannot share what they learn between similar words: \emph{heart attack}
and \emph{cardiac arrest} are two separate entries in the table, with nothing in
common.

\subsection{Neural language models}

The neural language model fixes exactly that. \citet{bengio2003neural} replace the
count table with a small neural network. Each word is first mapped to a vector, its
\emph{embedding}; the network reads the embeddings of the context and predicts the
next word with a softmax over the vocabulary. Because it learns the embeddings at
the same time as the prediction, words used in similar ways end up with similar
vectors and can pool their evidence. A sequence never seen in training is no longer
a dead end, because its words resemble others the model has seen. This alone cut
the perplexity of the best n-gram models by about a quarter. The embeddings turned
out to be useful on their own: word2vec \citep{mikolov-etal-2013-linguistic} showed
that simple arithmetic on them captures regularities of meaning; the textbook
example is that \emph{king} $-$ \emph{man} $+$ \emph{woman} lands near \emph{queen}.
These early vectors were still \emph{static}: one vector per word, whatever the
sentence. ELMo \citep{peters-etal-2018-deep} made them contextual, giving a word a
different vector in each sentence, and showed that representations pretrained on
plain text and then reused improve a wide range of tasks. This is an early form of
the pretrain-then-adapt recipe the rest of this thesis depends on.

\subsection{Recurrent networks}

All of these still read a fixed window of context. A recurrent network removes that
limit by processing a sequence one token at a time and carrying a hidden state
forward, so that in principle each prediction can depend on everything before it.
Recurrent language models \citep{mikolov10_interspeech} improved over n-grams, and
the LSTM \citep{hochreiter1997long} was designed to keep information over longer
spans. Two problems remained. The gradient that trains these models tends to vanish
over long sequences, so distant context is hard to learn, and the step-by-step
processing cannot be parallelised, which makes training on large corpora slow.
Long-range dependencies and parallelism are the two limits the next architecture
was built to solve.


\section{Transformer}
\label{sec:lm-transformer}

\subsection{Self-attention}

The Transformer \citep{vaswani2017attention} keeps the goal, predicting the next
token, and changes how context is read. Its core operation is \emph{self-attention},
which lets every token look directly at every other. From the input representations
$X$, the model computes three projections, a set of queries, keys, and values:
\[
  Q = XW_Q, \quad K = XW_K, \quad V = XW_V.
\]
Each token's query is compared with every key through a dot product, which measures
how relevant the other tokens are to it. These scores are turned into weights with a
softmax, and the token's new representation is the weighted average of the values:
\[
  \mathrm{Attention}(Q,K,V) = \mathrm{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V.
\]
The division by $\sqrt{d_k}$, the size of the key vectors, keeps the dot products
from growing too large, which would push the softmax to put almost all the weight on
a single token.

Two things follow. Any two tokens now interact in a single step, however far apart
they are, whereas a recurrent network had to pass information through every position
in between. And since the scores for all positions are computed together as matrix
products, the whole sequence is processed at once, which is what made training on
very large corpora practical. The price is cost: every token is compared with every
other, so time and memory grow with the square of the sequence length, $O(L^2)$. In
practice the operation runs in several \emph{heads} side by side, each with its own
projections, so that different heads can follow different things, such as syntax,
meaning, or position.

\subsection{Positional encodings}

Self-attention has one blind spot: it treats its input as a set, with no sense of
order. \emph{The cat sat} and \emph{sat the cat} look the same to it, so position
has to be supplied separately. The original Transformer added fixed sinusoidal
signals to the embeddings \citep{vaswani2017attention}, and BERT instead learned a
vector for each position \citep{devlin_bert_2019}; both encode an \emph{absolute}
position and cope badly with sequences longer than those seen in training. Later
methods encode \emph{relative} position. ALiBi adds a small penalty that grows with
the distance between two tokens \citep{alibi}, and RoPE rotates the queries and keys
by an angle that depends on their position, so that their dot product ends up
depending only on how far apart they are \citep{rope}. RoPE handles long and
variable lengths well and is now the default in recent models such as LLaMA,
Mistral, and ModernBERT.

\subsection{Encoders and decoders}

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\scriptsize\sffamily]
\def\s{0.52}
\foreach \name/\dx/\causal in {{Encoder (full)}/0/0, {Decoder (causal)}/4.4/1}{
  \begin{scope}[xshift=\dx cm]
    \node[font=\footnotesize\sffamily, text=ThesisInk] at (2*\s,0.92) {\name};
    \foreach \j/\t in {0/The,1/cat,2/sat,3/down}
      \node[text=ThesisNeutral] at (\j*\s+0.5*\s,0.42) {\t};
    \foreach \i/\t in {0/The,1/cat,2/sat,3/down}{
      \node[text=ThesisNeutral, anchor=east] at (-0.06,-\i*\s-0.5*\s) {\t};
      \foreach \j in {0,1,2,3}{
        \draw[draw=ThesisInk!15, line width=0.3pt]
          (\j*\s,-\i*\s) rectangle (\j*\s+\s,-\i*\s-\s);
        \pgfmathtruncatemacro{\on}{(\causal==0) || (\j<=\i)}
        \ifnum\on=1
          \fill[ThesisPrimary!70]
            (\j*\s+0.06,-\i*\s-0.06) rectangle (\j*\s+\s-0.06,-\i*\s-\s+0.06);
        \fi
      }
    }
    \node[text=ThesisNeutral, font=\tiny] at (2*\s,-4.6*\s) {key $j\rightarrow$};
    \node[text=ThesisNeutral, font=\tiny, rotate=90, anchor=south]
      at (-0.62,-2*\s) {query $i\downarrow$};
  \end{scope}
}
\end{tikzpicture}
\caption{Attention masks as matrices. Row $i$ is a query token and column $j$ a key
token; a filled cell means token $i$ may attend to token $j$. The encoder allows
every pair, so the matrix is full; the decoder allows only $j \le i$, the lower
triangle, so each token sees itself and the tokens before it but none of those that
follow.}
\label{fig:lm-attention}
\end{figure}

The same Transformer block can be wired in two ways, and the only difference is
which tokens each one is allowed to see (\Cref{fig:lm-attention}). In an \emph{encoder}, every token attends
to the whole sequence, left and right. This two-sided view suits tasks that read a
text and label it, such as classification or entity recognition, and it underlies
BERT \citep{devlin_bert_2019} and RoBERTa \citep{roberta}. In a \emph{decoder}, a
token may attend only to the tokens before it, a restriction imposed by a causal
mask. This is what makes generation possible, one token at a time, as in GPT
\citep{Radford2018ImprovingLU} and LLaMA. A third design keeps both: an encoder to
read an input and a decoder to write an output, as in T5 \citep{raffel2020t5},
which is natural for translation and other input-to-output tasks but less common
today. The two are built the same way. What really separates them is the training
objective, the subject of the next section.


\section{Pretraining Objectives}
\label{sec:lm-objectives}

The architecture fixes how a model reads its context; the pretraining objective
fixes what it is asked to predict, and that choice shapes the representations it
learns. Two objectives dominate, and they differ in a way that matters for the rest
of this thesis (\Cref{fig:lm-objectives}).

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[font=\scriptsize\sffamily,
  cell/.style={draw=ThesisInk!22, rounded corners=1.5pt, minimum width=1.25cm,
    minimum height=0.55cm, text=ThesisInk, inner sep=2pt},
  on/.style={fill=ThesisPrimary!30, draw=ThesisPrimaryDark},
  msk/.style={fill=ThesisSecondary!38, draw=ThesisSecondaryDark},
  tgt/.style={rounded corners=1.5pt, minimum width=1.25cm, minimum height=0.5cm,
    text=ThesisInk, fill=ThesisPrimary!16, draw=ThesisPrimaryDark!60, inner sep=2pt},
  tgtm/.style={tgt, fill=ThesisSecondary!22, draw=ThesisSecondaryDark!70},
  up/.style={->, ThesisInk!45, shorten >=1.5pt, shorten <=1.5pt},
]
\def\sx{1.6}
% ===== Causal language modeling =====
\node[font=\footnotesize\sffamily\bfseries, text=ThesisInk, anchor=west] at (-0.85,1.95)
  {Causal language modeling};
\node[font=\scriptsize, text=ThesisNeutral, anchor=east] at (-0.78,1.0) {target};
\node[font=\scriptsize, text=ThesisNeutral, anchor=east] at (-0.78,0.0) {input};
\foreach \i/\w/\n in {0/The/cat,1/cat/sat,2/sat/on,3/on/the,4/the/mat,5/mat/{$\langle$eos$\rangle$}}{
  \node[cell,on] (c\i) at (\i*\sx,0) {\w};
  \node[tgt] (t\i) at (\i*\sx,1.0) {\n};
  \draw[up] (c\i) -- (t\i);
}
% ===== Masked language modeling =====
\begin{scope}[yshift=-2.95cm]
\node[font=\footnotesize\sffamily\bfseries, text=ThesisInk, anchor=west] at (-0.85,1.95)
  {Masked language modeling};
\node[font=\scriptsize, text=ThesisNeutral, anchor=east] at (-0.78,1.0) {target};
\node[font=\scriptsize, text=ThesisNeutral, anchor=east] at (-0.78,0.0) {input};
\foreach \i/\w/\n/\m in {0/The//0,1/{[MASK]}/cat/1,2/sat//0,3/on//0,4/{[MASK]}/the/1,5/mat//0}{
  \ifnum\m=1
    \node[cell,msk] (m\i) at (\i*\sx,0) {\w};
    \node[tgtm] (s\i) at (\i*\sx,1.0) {\n};
    \draw[up] (m\i) -- (s\i);
  \else
    \node[cell] (m\i) at (\i*\sx,0) {\w};
  \fi
}
\end{scope}
\end{tikzpicture}}
\caption{The same sentence under the two objectives. The bottom row is the input
and the boxes above are the targets the model must predict, one arrow per predicted
position. Causal language modeling predicts the next token at every position, so
every token gives a training signal (dense). Masked language modeling hides about
15\% of the tokens, here two, and predicts only those from the context on both
sides, so most positions give no signal (sparse). This gap in how much signal each
token carries is what the encoder pretraining of \Cref{chap:objectives} exploits.}
\label{fig:lm-objectives}
\end{figure}

\subsection{Causal language modeling}

The most direct objective follows the chain rule: predict the next token. The model
reads the sequence once and, at each position $t$, produces a probability
distribution over the vocabulary for the token that should come next. Training
lowers the negative log-probability it assigns to the token that actually follows,
summed over the whole sequence,
\[
  \mathcal{L}_{\text{CLM}} = -\sum_{t=1}^{L} \log P(w_t \mid w_{<t}),
\]
where $w_{<t}$ is the tokens before position $t$. A causal mask is what lets this
happen in a single pass: each position may look only to its left, so the model can
be scored at every position at once without seeing the very token it must predict.
Every position therefore yields a prediction and a gradient, so the supervision is
dense. This is the objective behind GPT \citep{Radford2018ImprovingLU} and the
modern decoder models, and it fits generation naturally: at inference the model does
exactly what it was trained to do, continue the text.

\subsection{Masked language modeling}

Understanding a text, rather than continuing it, calls for a different objective.
BERT \citep{devlin_bert_2019} corrupts the input by replacing some tokens with a
special \texttt{[MASK]} symbol, then trains the model to put the originals back. At
each masked position the model predicts the missing token from the context on both
sides, and the loss counts only those positions,
\[
  \mathcal{L}_{\text{MLM}} = -\sum_{t \in \mathcal{M}} \log P(w_t \mid \mathbf{w}_{\setminus t}),
\]
where $\mathcal{M}$ is the set of masked positions and $\mathbf{w}_{\setminus t}$ is
the rest of the (corrupted) sequence. About 15\% of the tokens are chosen; of those,
80\% become \texttt{[MASK]}, 10\% are swapped for a random word, and 10\% are left
as they are, a mix that stops the model from only ever reacting to the
\texttt{[MASK]} symbol. Seeing the whole sequence at once makes the representations
bidirectional, which is what makes encoders strong at classification and entity
recognition. The price is that only the masked positions produce a gradient, about
one token in seven, against every token under the causal objective.

\subsection{Mixing the two}

The two objectives trade the same quantity against itself: causal training gives a
signal at every token but only looks left, while masked training looks both ways but
learns from few tokens. Recent work tries to get both at once. Some encoders are
pretrained with a causal phase followed by a short masked phase, which beats masked
pretraining alone \citep{gisserot2025clm}; others alternate the two objectives during
training \citep{antlm2024}. This line is central to the thesis: \Cref{chap:objectives}
shows that a temporary detour through causal pretraining leaves a lasting and useful
mark on an encoder. An earlier idea, ELECTRA's replaced-token detection
\citep{electra}, also sought a denser signal, by training the model to spot tokens a
small generator had swapped in, though recent encoders such as ModernBERT
\citep{modernbert} show that a good architecture with plain masking is already
enough.

With both the architecture and the objective settled, the next question is how far
this recipe scales.


\section{Scaling Laws \& LLMs}
\label{sec:lm-scaling}

\subsection{Scaling}

\begin{itemize}
  \item GPT-3 (Brown et al., 2020): 175B params, few-shot emergent, prompting remplace le finetuning
  \item Chinchilla (Hoffmann et al., 2022):
    $$
    L(N,D) = \frac{A}{N^\alpha} + \frac{B}{D^\beta} + E
    $$
    -> compute-optimal: smaller model + more data bat larger model + less data
  \item LLaMA (Touvron et al., 2023): 7B rivalise avec bcp plus gros via 1T tokens
    -> open weights, democratisation
\end{itemize}

\subsection{Post-ChatGPT}

\begin{itemize}
  \item RLHF (Ouyang et al., 2022) -> preference learning
  \item Instruction tuning
  \item ChatGPT, Claude, Gemini
  \item Emergent abilities (Wei et al., 2022) -> chain-of-thought etc.
\end{itemize}

% TODO section courte, juste poser le contexte

-> ok ca scale pour le general, mais nous on a un domaine specifique -> comment adapter?


\section{Continual Pretraining}
\label{sec:lm-continual}

Section centrale pour la these. C'est le coeur de ce qu'on fait (CamemBERT-bio, ModernCamemBERT-bio).

\subsection{Fondements et terminologie}

``Don't Stop Pretraining'' (Gururangan et al., ACL 2020):
\begin{itemize}
  \item DAPT = Domain-Adaptive Pre-Training: 2eme phase sur corpus domaine
  \item TAPT = Task-Adaptive Pre-Training: sur donnees non-annotees de la tache
  \item DAPT + TAPT combine -> meilleurs resultats
  \item Jusqu'a +3 F1 meme avec petit corpus
\end{itemize}

Taxonomie moderne (Wang et al., CSUR 2025):
\begin{itemize}
  \item CPT = nouvelles donnees generales au fil du temps
  \item DAP = specialisation domaine
  \item CFT = adaptation sequentielle de taches
\end{itemize}

\subsection{Catastrophic Forgetting}

Le probleme (Kirkpatrick et al., 2017):
\begin{itemize}
  \item Les reseaux ecrasent les poids necessaires aux taches precedentes
  \item EWC (Elastic Weight Consolidation): penaliser les changements sur les poids importants
\end{itemize}

Pourquoi ca compte pour les LLMs:
\begin{itemize}
  \item Continual PT peut degrader les capacites generales
  \item Capacite a suivre des instructions particulierement fragile
  \item Petits modeles plus sensibles (Yildiz et al., 2024)
\end{itemize}

``Spurious forgetting'' (2024): les baisses de perf refletent peut-etre une perte d'alignement, pas une vraie perte de connaissance.

-> ok c'est un vrai probleme, quelles solutions?

\subsection{Solutions recentes (2024-2025)}

Ibrahim et al. (2024), ``Simple and Scalable Strategies'':
\begin{itemize}
  \item LR re-warming: relancer le learning rate quand on ajoute des donnees
  \item LR re-decaying: cosine decay apres warmup
  \item Replay: mixer 1\% des donnees precedentes -> ca suffit
  \item Valide a 405M et 10B
  \item -> egalise from-scratch pour une fraction du compute
\end{itemize}

Stability gap (arXiv:2406.14833, 2024):
\begin{itemize}
  \item Chute temporaire au debut du continual PT, puis recovery
  \item Solutions: subset propre, sous-corpus de qualite, data mixing
  \item 36.2\% -> 40.7\% avec seulement 40\% du budget training
\end{itemize}

Data selection:
\begin{itemize}
  \item Donnees non pertinentes = degradation
  \item 10\% du corpus bien selectionne fait aussi bien que 100\% vanilla continual PT
\end{itemize}

\subsection{From scratch vs continual PT}

\begin{itemize}
  \item From scratch (PubMedBERT, DrBERT): vocab custom, pas de forgetting, mais tres cher
  \item Continual (BioBERT, Gururangan): efficient, preserve le general, mais tokenizer sous-optimal
\end{itemize}

Cout compute:
\begin{itemize}
  \item DrBERT from scratch: 128x V100, 20h
  \item AliBERT from scratch: 48x A100, 20h
  \item BioBERT continual: 8x V100, \~{}10 jours
  \item -> from-scratch coute \~{}30x plus de compute
\end{itemize}

% TODO tableau

\subsection{Encoders vs decoders: dynamiques differentes}

Decoders (LLMs), ca marche pas bien:
\begin{itemize}
  \item BioMistral: $-$0.9 points apres continual PT
  \item Dorfner et al. (2024): biomedical LLMs sous-performent les generalistes
  \item OpenBioLLM-8B: 30\% vs Llama-3-8B: 64.3\% sur cas NEJM
  \item -> hypothese: perte de connaissances generales pendant la specialisation
\end{itemize}

Encoders, c'est different:
\begin{itemize}
  \item Contexte bidirectionnel peut-etre plus robuste au forgetting
  \item MLM moins sujet au catastrophic forgetting?
  \item -> voir Ch.5: approche CLM vers MLM sur les encoders
\end{itemize}

-> tension narrative: continual PT marche pas pour les decoders, et pour les encoders? -> c'est la question de la these

\subsection{Tentatives anterieures en francais biomedical}

\begin{itemize}
  \item Copara et al. (2020): 31K articles -> aucune amelioration (corpus trop petit)
  \item Le Clercq de Lannoy et al. (2022): 136M mots -> +2 EMEA seulement
  \item Dura et al. (2022): 21M docs APHP -> +3\% mais donnees privees
\end{itemize}

Question ouverte: pourquoi des resultats contradictoires? DrBERT dit que continual PT marche pas. CamemBERT-bio (cette these) montre que si, avec le bon corpus.

-> ok on sait ce qu'est le continual PT, maintenant quels modeles existent?


\section{Modeles pretrained biomed}
\label{sec:lm-biomedical}

\subsection{Encoders biomed anglais}

\begin{itemize}
  \item BioBERT (Lee et al., 2019): PubMed+PMC 18B mots, continual -> +0.62\% NER, +2.80\% RE
  \item SciBERT (Beltagy et al., 2019): Semantic Scholar 1.14M papers, from scratch -> 42\% vocab overlap avec BERT
  \item PubMedBERT (Gu et al., 2022): PubMed only, from scratch -> introduit BLURB benchmark
  \item ClinicalBERT (Alsentzer et al., 2019): MIMIC-III, continual -> acces restreint
  \item BioLinkBERT (Yasunaga et al., 2022): PubMed + citation links, from scratch -> +7\% BioASQ
\end{itemize}

SOTA 2025: BioClinical ModernBERT, cf section architectures modernes.

% TODO tableau

\subsection{Modeles francais}

General:
\begin{itemize}
  \item CamemBERT (Martin et al., 2020): RoBERTa sur OSCAR 138GB
    -> resultat interessant: 4GB donne a peu pres la meme perf que 138GB
  \item FlauBERT (Le et al., 2020)
  \item CamemBERT 2.0 / CamemBERTa (2024): architecture DeBERTaV3, nouveau tokenizer
\end{itemize}

Biomedical francais:
\begin{itemize}
  \item DrBERT (Labrak et al., 2023): from scratch, public + prive
    -> pretend que continual PT marche pas pour le francais biomed
  \item AliBERT (Berhe et al., 2023): from scratch, Unigram regularise
    -> pas disponible publiquement
\end{itemize}

-> tension: DrBERT dit from-scratch necessaire, CamemBERT-bio (cette these) montre le contraire

\subsection{Decoders biomed}

\begin{itemize}
  \item BioMistral (Labrak et al., 2024): 3B tokens PMC, 32x A100 20h, base Mistral
  \item Meditron (Chen et al., 2023): 46B tokens, 128x A100 332h, base Llama-2
  \item PMC-LLaMA (Wu et al., 2024): 75B tokens, massif
\end{itemize}

Continual PT pour decoders = resultats mitiges:
\begin{itemize}
  \item BioMistral: $-$0.9 points, recupere seulement via model merging
  \item Dorfner et al. (2024): ``Biomedical LLMs Seem not to be Superior to Generalist Models''
  \item -> contraste avec les encoders, voir Ch.5
\end{itemize}

% TODO tableau

-> les modeles existent mais les archis ont evolue, quoi de neuf?


\section{Architectures modernes}
\label{sec:lm-modern}

\subsection{Modernisation des encoders}

DeBERTa (He et al., ICLR 2021):
\begin{itemize}
  \item Disentangled attention: separe contenu et position
  \item Premier a battre l'humain sur SuperGLUE
  \item -> influence sur ModernBERT, CamemBERT 2.0
\end{itemize}

MosaicBERT (Portes et al., NeurIPS 2023):
\begin{itemize}
  \item 30\% masking au lieu de 15\% BERT
  \item FlashAttention + ALiBi + GLU + unpadding
  \item BERT-base en 1.13h sur 8x A100 (\~{}\$20)
  \item -> fondation directe de ModernBERT
\end{itemize}

\subsection{Efficient Attention}

FlashAttention (Dao et al., 2022):
\begin{itemize}
  \item Reorganise le calcul pour la hierarchie memoire GPU (tiling, kernel fusion)
  \item Meme $O(L^2)$ mais 2-4x plus rapide en pratique
  \item Standard maintenant
\end{itemize}

Autres:
\begin{itemize}
  \item GQA (Ainslie et al., 2023): partage de KV-heads -> Llama-2, Mistral
  \item RMSNorm (Zhang \& Sennrich, 2019): LayerNorm simplifie
\end{itemize}

\subsection{Long context}

Le probleme: BERT = 512 tokens, comptes-rendus cliniques = souvent 2000+. Tronquer = perdre de l'info diagnostique.

\begin{itemize}
  \item Longformer (Beltagy et al., 2020): 4096 tokens, sliding window + global attention
  \item BigBird (Zaheer et al., 2020): 4096, sparse attention
  \item Clinical Longformer (Li et al., 2022): 4096, Longformer continue pretrained sur MIMIC-III
  \item ModernBERT (Warner et al., 2024): 8192 tokens
    \begin{itemize}
      \item FlashAttention v2/v3
      \item RoPE
      \item Alternance local/global attention
      \item 30\% masking (de MosaicBERT)
      \item -> 16x le contexte de BERT
    \end{itemize}
\end{itemize}

Adaptations cliniques 2025:
\begin{itemize}
  \item Clinical ModernBERT (Lee et al., 2025): PubMed + MIMIC-IV + ontologies
  \item BioClinical ModernBERT (2025): 53.5B tokens, 20 datasets cliniques -> SOTA actuel
\end{itemize}

% TODO tableau


\section{Tokenization}
\label{sec:lm-tokenization}

\subsection{Methodes subword}

\begin{itemize}
  \item BPE (Sennrich et al., 2016): merge les paires les plus frequentes
  \item SentencePiece (Kudo \& Richardson, 2018): language-independent
  \item Unigram (Kudo, 2018): selection probabiliste
\end{itemize}

\subsection{Domain mismatch}

SciBERT: vocab general vs scientifique = 42\% intersection seulement.

Exemple biomed francais:
\begin{itemize}
  \item Tokenizer general: ``echocardiographie'' -> [``echo'', ``\#\#cardi'', ``\#\#ographie'']
  \item Tokenizer domaine: ``echocardiographie'' -> [``echocardiographi'', ``e'']
\end{itemize}

Tradeoff:
\begin{itemize}
  \item From-scratch = vocab custom mais cher
  \item Continual PT = tokenizer sous-optimal mais efficient
\end{itemize}


\section{Limites et transition}
\label{sec:lm-limitations}

\begin{itemize}
  \item Rarete des donnees: texte clinique prive (CNIL), modeles hospitaliers pas partageables
    -> besoin d'alternatives publiques -> Ch.3 Corpus Annotation
  \item Knowledge gap: LMs captent des patterns, pas la structure des ontologies. Codage medical = connaissances hierarchiques
    -> approches knowledge-enhanced -> Ch.4 Clinical IE
  \item Documents longs: comptes-rendus hospitaliers depassent les limites de BERT. Longformer et ModernBERT aident mais faut adapter au domaine
\end{itemize}

-> ce chapitre a couvert les fondations techniques du language modeling. Le chapitre suivant traite de comment construire et annoter des corpus pour entrainer ces modeles.

% TODO rediger la transition proprement
