\partepigraph{``the speaking of language is part of an activity, or of a form of
life''}{%
Ludwig Wittgenstein, \textit{Philosophical Investigations}}

The first two parts built a public biomedical corpus and pretrained encoders on
it. This part uses one of those encoders on a clinical task.

The task comes from clinical research. A study follows each of its patients over
time and keeps a structured form for each one, with a fixed set of fields such as
the treatment given or the date it started. The values are already written down,
in the consultation notes, discharge summaries and pathology results produced
during care. Filling the form means reading those reports and putting each value
in the right field. The difficulty is that the reports are written at successive
moments as the disease unfolds, so a value stated in one report can be corrected
in a later one, and the model has to work out which statement still holds.

The same constraint applies here as in the first two parts. There is no corpus of
French hospital reports paired with the form a clinician filled from them, so the
supervision has to be built rather than collected. Confidentiality also keeps
those records away from a cloud API, which means the model has to run inside the
hospital, on the hardware a clinical service has. The next three chapters build
the supervision (\Cref{chap:synthetic}), design an extractor that can be asked
for new fields at inference time (\Cref{chap:architectures}), and evaluate it
against much larger generative models (\Cref{chap:evaluation}). The question they
share is whether public and generated text are enough for a task whose real data
no one can show us.

\chapter{Synthetic Data for Task Adaptation}
\label{chap:synthetic}
\input{sources/part_3/chapter9/article}

\chapter{Architectures for Low-Resource Extraction}
\label{chap:architectures}
\input{sources/part_3/chapter8/article}

\chapter{Evaluating Open-Vocabulary Extraction}
\label{chap:evaluation}
\input{sources/part_3/chapter7/article}
