% This chapter is based on:
%   Rian Touchent et al.
%   The CLM Detour: Computational Hysteresis Improves
%   Encoder Continued Pretraining.
%   COLM, 2026.

\section{Introduction}

Masked language modeling sees only 15\% of its input. At every training step, 85\% of the tokens receive no gradient signal. Causal language modeling, by contrast, predicts every token from its predecessors: every position contributes, every position learns. The price is that it can only look backward. The question is whether an encoder can benefit from this dense training signal without permanently losing its ability to look in both directions.

In this chapter, we show that it can. We introduce a two-phase pretraining strategy for ModernCamemBERT-bio: first, continual pretraining with a causal language modeling objective, then a short ``decay'' phase that converts the model back to masked language modeling. The resulting encoder outperforms the standard MLM-only approach by +2.9 $F_1$ on average across 8 French biomedical tasks (Base) and +1.5 $F_1$ (Large), winning on all 8. The effect transfers to English at three scales, with gains of +0.3 to +0.8 $F_1$ depending on model size.

The CLM phase leaves a lasting imprint on low transformer layers. CKA analysis shows that layers 0--7 diverge 5--44$\times$ more between CLM and MLM models than seed noise alone, and this divergence persists through extended MLM decay. Freeze interventions establish a double dissociation: freezing low layers during CLM eliminates the downstream benefit; freezing mid layers preserves it. The advantage cannot be localized post hoc: transplanting layer groups from CLM to MLM recovers at most 35\% of the gap. The effect is emergent, requiring the coordinated interaction between all layers.


\section{Motivation}

CamemBERT-bio (\Cref{chap:encoders}) demonstrated that continual pretraining works for biomedical encoders. However, it used the original CamemBERT architecture: 512-token context, absolute position embeddings, no Flash Attention. Clinical documents are often longer than 512 tokens, and the architecture has not kept pace with recent advances.

ModernBERT~\citep{modernbert} addresses these limitations with Flash Attention, RoPE position embeddings, alternating local/global attention, and an 8,192-token context window. For biomedical adaptation, we start from ModernCamemBERT (French) and ModernBERT (English), in both Base (150M) and Large (350M) sizes.

The standard approach would be to continue MLM pretraining on biomedical data, as we did for CamemBERT-bio. But MLM has a structural limitation: only 15\% of tokens are masked and receive gradient signal at each step. Causal language modeling does not have this limitation: every token predicts the next, and every position contributes to the loss. The idea is to use CLM for the compute-intensive pretraining phase, and switch back to MLM only at the end, to recover bidirectional attention. The key question is whether this switch erases the benefits of CLM or preserves them.

\paragraph{Related approaches.} A handful of recent works explore objective switching for encoder pretraining. \citet{gisserot2025clm} pretrain encoders \emph{from scratch} (210M--1B parameters, 100B tokens) and find that a biphasic CLM-then-MLM schedule outperforms pure MLM under fixed compute. \citet{ettin2025} train matched encoder/decoder pairs (up to 1B parameters, 1.7T tokens) and show that continued pretraining on the reverse objective does not bridge the encoder-decoder performance gap. AntLM~\citep{antlm2024} alternates between CLM and MLM epochs at small scale (10M words). Our approach differs in three ways: we study the round-trip MLM$\to$CLM$\to$MLM with continued pretraining (not from scratch); we focus on the lasting effect of the CLM phase after returning to MLM; and we provide causal evidence via freeze interventions of where the benefit comes from. As \citet{gisserot2025clm} note, understanding how training objectives transfer to downstream tasks ``remains largely unexplored.''


\section{Method}
\label{sec:clm-detour}

\subsection{The CLM Detour}

Our approach consists of two phases, illustrated in \Cref{fig:clm-pipeline}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=1.8cm, >=Stealth, scale=0.85, transform shape,
  box/.style={rectangle, rounded corners=3pt, draw=ThesisNeutral, minimum height=0.9cm, minimum width=2.4cm, align=center, font=\small\sffamily}
]
  \node[box, fill=ThesisPrimary!20] (base) {MLM encoder\\(ModernBERT)};
  \node[box, fill=ThesisTertiary!30, right=of base] (clm) {CLM phase\\(10--50B tokens)};
  \node[box, fill=ThesisSecondary!30, right=of clm] (decay) {MLM decay\\(10\% of Phase 1)};
  \node[box, fill=ThesisPrimary!30, right=of decay] (out) {ModernCamem-\\BERT-bio};

  \draw[->, thick, ThesisNeutral] (base) -- (clm) node[midway, above, font=\tiny\sffamily] {causal mask};
  \draw[->, thick, ThesisNeutral] (clm) -- (decay) node[midway, above, font=\tiny\sffamily] {bidir.\ mask};
  \draw[->, thick, ThesisNeutral] (decay) -- (out);
\end{tikzpicture}
\caption{\textbf{The CLM detour pipeline.} Starting from a pretrained MLM encoder, we continue pretraining with a causal objective (Phase~1), then decay back to MLM (Phase~2). The model does not return to an MLM-like state.}
\label{fig:clm-pipeline}
\end{figure}

\paragraph{Phase 1: CLM pretraining.} Starting from a pretrained ModernBERT encoder, we replace the bidirectional attention mask with a causal mask and train with next-token prediction. The model architecture is unchanged: only the attention mask and loss computation differ. This forces the model to compress information into each token representation, since it cannot look ahead.

\paragraph{Phase 2: MLM decay.} After CLM pretraining, we restore bidirectional attention and train with MLM at 15\% masking for approximately 10\% of the Phase~1 budget. The optimizer state is kept between phases; only the learning rate scheduler resets. Phase~2 decays the learning rate from peak to 10\% of peak.

\paragraph{Controlled comparison.} The MLM baseline follows the same two-phase structure with 30\% masking in Phase~1 and 15\% in Phase~2, identical schedule and optimizer. Both trajectories use identical architecture, data, and total compute:
\begin{align*}
\theta_A &= T_{\text{MLM}}(T_{\text{CLM}}(\theta_0, n), 0.1n) && \text{(CLM detour)} \\
\theta_B &= T_{\text{MLM}}(\theta_0, 1.1n) && \text{(MLM baseline)}
\end{align*}
The only difference is the Phase~1 objective. Any performance gap is attributable to the training objective alone.

\subsection{Freeze Interventions}
\label{sec:freeze}

To test which layers carry the CLM benefit, we run three freeze experiments on the 22-layer French Base model, where the CLM-MLM gap is largest (+2.9pp):

\begin{itemize}
    \item \textbf{Experiment~1} (low freeze, CLM phase): Layers 0--7 frozen during CLM, then normal decay. Tests whether low-layer modification during CLM is necessary.
    \item \textbf{Experiment~2} (low freeze, decay phase): Normal CLM, then layers 0--7 frozen during decay. Tests whether low-layer CLM changes persist without further updates.
    \item \textbf{Experiment~3} (mid freeze, CLM phase): Layers 8--14 frozen during CLM, then normal decay. Together with Experiment~1, tests selectivity: if freezing low layers eliminates the benefit while freezing mid layers preserves it, this constitutes a double dissociation.
\end{itemize}

\subsection{Data}

\paragraph{French.} We compile 10B tokens (Base) or 25B tokens (Large) from four sources. The main source (7B tokens) is French biomedical literature (scientific articles, clinical guidelines, medical theses), where each paragraph is scored for educational value using Qwen3-235B and articles are upsampled based on their proportion of high-scoring paragraphs, following the Biomed-Enriched approach (\Cref{chap:quality}). The remaining sources are synthetic medical QA from French coding systems (2B), clinical cases from the European Clinical Case Corpus (E3C, 400M), and drug package inserts from the European Medicines Agency (600M).

\paragraph{English.} For the 50B scale, we mix three components: 60\% biomedical literature from Biomed-Enriched (PMC Open Access articles filtered by educational value $\geq 3.0$), 20\% medical instruction-following datasets (MedS-Ins, ReasonMed, MediFlow, UltraMedical, EHR-Ins-Reasoning), and 20\% MIMIC-III discharge summaries and clinical notes. The smaller 10B variant uses Biomed-Enriched with clinical case paragraphs upsampled 100$\times$ and other clinical documents 10$\times$ (80\%), plus MediFlow medical instructions (20\%), without MIMIC.

\subsection{Training Details}

All runs use decoupled AdamW with peak lr $2 \times 10^{-4}$, $\beta_1 = 0.9$, $\beta_2 = 0.98$, weight decay $10^{-5}$, and a global batch size of 384 sequences ($\sim$3.1M tokens). Phase~1 uses linear warmup over 100M tokens then constant learning rate. Documents are packed into 8,192-token sequences with end-of-sequence tokens between documents. Training uses bf16 mixed precision on 4$\times$H100 GPUs.

\subsection{Fine-tuning Protocol}

\paragraph{French tasks.} Classification (DiaMED) and multilabel tasks (MedDialog, FrACCO, CANTEMIST, DisTEMIST) use a linear head on the CLS token, trained for 15 epochs with lr $= 2 \times 10^{-5}$, batch size 4, and 10\% warmup. Multilabel tasks use max sequence length 4096 with weighted BCEWithLogitsLoss; classification uses 2048. NER tasks (EMEA, Medline) use a span-based architecture with BiLSTM, CRF, and biaffine scorer, trained for 4000 steps with lr $= 5 \times 10^{-5}$ for the encoder and $10^{-3}$ for the head, batch size 16, max length 1024. All tasks use AdamW with weight decay 0.01 and select the best checkpoint by validation $F_1$.

\paragraph{English tasks.} We follow the BioClinical-ModernBERT evaluation protocol~\citep{sounack2025bioclinicalmodernbert}: lr $= 5 \times 10^{-5}$, weight decay 0.01, batch size 16, 10 epochs for most tasks (20 for NER). All models are fine-tuned with the same hyperparameters per task.

\subsection{Evaluation}

We evaluate on 8 French and 11 English biomedical tasks (\Cref{tab:eval-tasks}), using 9 seeds for French and 5 for English. French baselines include ModernCamemBERT, DrBERT~\citep{labrak2023drbert}, CamemBERT-bio, and CamemBERT~\citep{martin-etal-2020-camembert}. English baselines include ModernBERT, PubMedBERT~\citep{gu_domain-specific_2022}, BioBERT~\citep{lee_biobert_2019}, and BioClinical-ModernBERT~\citep{sounack2025bioclinicalmodernbert}.

\begin{table}[t]
\centering
\small
\caption{\textbf{Evaluation tasks.} Max sequence length reflects the fine-tuning configuration; longer values indicate tasks with longer input documents.}
\label{tab:eval-tasks}
\begin{tabular}{llcc}
\toprule
\textbf{Task} & \textbf{Type} & \textbf{Lang} & \textbf{Max len.} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{French}} \\[1pt]
DiaMED & Classification & FR & 2048 \\
MedDialog-FR & Multilabel & FR & 4096 \\
FrACCO (30/100) & Multilabel & FR & 4096 \\
CANTEMIST & Multilabel & FR & 4096 \\
DisTEMIST & Multilabel & FR & 4096 \\
QUAERO (EMEA/Medline) & NER & FR & 1024 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{English}} \\[1pt]
ChemProt & Relation Extr. & EN & 512 \\
Phenotyping & Classification & EN & 8192 \\
COS & NER & EN & 512 \\
Social History & NER & EN & 512 \\
De-identification & NER & EN & 8192 \\
AnatEM & NER & EN & 512 \\
BC5CDR & NER & EN & 512 \\
JNLPBA & NER & EN & 512 \\
NCBI Disease & NER & EN & 512 \\
GAD & Classification & EN & 512 \\
HoC & Classification & EN & 512 \\
\bottomrule
\end{tabular}
\end{table}


\section{Results}

\subsection{French Biomedical Tasks}

\begin{table*}[t]
\centering
\small
\caption{\textbf{French biomedical results} (macro $F_1$, 9 seeds). Bold: best per column; underline: second best.}
\label{tab:french-results}
\resizebox{\textwidth}{!}{
\begin{tabular}{lcccccccccc}
\toprule
& & \multicolumn{5}{c}{\textbf{Multilabel Classification}} & {\textbf{CLS}} & \multicolumn{2}{c}{\textbf{NER}} & \\
\cmidrule(lr){3-7} \cmidrule(lr){8-8} \cmidrule(lr){9-10}
& \textbf{Ctx} & \textbf{FrACCO-30} & \textbf{FrACCO-100} & \textbf{CANTEMIST} & \textbf{DISTEMIST} & \textbf{MedDialog} & \textbf{DiaMED} & \textbf{EMEA} & \textbf{Medline} & \textbf{Avg} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{11}{c}{\rule{0pt}{10pt}\textbf{Baselines}} \\[1pt]
ModernCamemBERT & 8192 & 70.1 & 55.3 & 63.3 & 20.2 & 60.6 & 56.4 & 63.4 & 55.3 & 55.6 \\
DrBERT & 512 & 53.0 & 35.6 & 37.9 & 21.4 & \underline{63.6} & 57.0 & \textbf{68.0} & \textbf{62.3} & 49.9 \\
CamemBERT-bio & 512 & 41.9 & 20.1 & 12.8 & 9.6 & 38.6 & 47.7 & 61.6 & 56.6 & 36.1 \\
CamemBERT & 512 & 40.8 & 19.4 & 11.9 & 9.5 & 37.4 & 40.6 & 61.5 & 54.7 & 34.5 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{11}{c}{\rule{0pt}{10pt}\textbf{Our Models (Base, 150M)}} \\[1pt]
MLM baseline & 8192 & 69.9 & 56.8 & 64.9 & 23.5 & 62.5 & 63.4 & 65.4 & 56.8 & 57.9 \\
CLM detour & 8192 & 74.8 & 60.1 & 71.0 & 25.5 & \underline{63.6} & \textbf{67.4} & 65.9 & 58.2 & 60.8 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{11}{c}{\rule{0pt}{10pt}\textbf{Our Models (Large, 350M)}} \\[1pt]
MLM baseline & 8192 & \underline{79.4} & \underline{63.3} & \underline{72.6} & \underline{29.1} & \textbf{64.5} & 61.2 & 66.1 & 58.0 & \underline{61.8} \\
CLM detour & 8192 & \textbf{80.7} & \textbf{65.4} & \textbf{74.4} & \textbf{30.4} & \textbf{64.5} & \underline{64.8} & \underline{67.0} & \underline{59.5} & \textbf{63.3} \\
\bottomrule
\end{tabular}
}
\end{table*}

\Cref{tab:french-results} presents the French results. For Base, the CLM detour achieves 60.8\% average $F_1$, outperforming the MLM baseline on all 8 tasks (+2.9pp). For Large, CLM reaches 63.3\% versus 61.8\% for MLM (+1.5pp). CamemBERT-bio and CamemBERT average 36\% and 35\% overall, limited by their 512-token context on long clinical documents. We release the CLM-detour models as ModernCamemBERT-bio (Base and Large).

\subsection{English Biomedical Tasks}

\begin{table*}[t]
\centering
\small
\caption{\textbf{English biomedical results} across 11 benchmarks (5 seeds). Base at 10B and 50B scales; Large at 50B. Bold: best; underline: second best.}
\label{tab:english-results}
\resizebox{\textwidth}{!}{
\sisetup{detect-weight=true}
\begin{tabular}{ll S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1] S[table-format=2.1]}
\toprule
& & \multicolumn{5}{c}{\textbf{Clinical Tasks}} & \multicolumn{6}{c}{\textbf{BigBIO Tasks}} & \\
\cmidrule(lr){3-7} \cmidrule(lr){8-13}
& \textbf{Ctx} & {\textbf{ChemPr}} & {\textbf{Pheno}} & {\textbf{COS}} & {\textbf{Social}} & {\textbf{DEID}} & {\textbf{AnatEM}} & {\textbf{BC5CDR}} & {\textbf{JNLPBA}} & {\textbf{NCBI}} & {\textbf{GAD}} & {\textbf{HoC}} & {\textbf{Avg}} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{14}{c}{\rule{0pt}{10pt}\textbf{Baselines}} \\[1pt]
ModernBERT-base & 8192 & 89.5 & 48.4 & 94.0 & 53.1 & 78.3 & 77.2 & 87.9 & 74.3 & 77.7 & 76.8 & 66.6 & 74.9 \\
PubMedBERT & 512 & 90.2 & 52.0 & 95.0 & 48.7 & 80.4 & \bfseries 83.3 & \underline{89.7} & 74.9 & \bfseries 82.1 & \underline{79.3} & \bfseries 71.0 & 77.0 \\
BioClinical-MdnBERT & 8192 & 90.0 & 60.7 & 94.8 & \underline{56.0} & 81.8 & 79.2 & 88.7 & 74.8 & 78.7 & 75.8 & 67.0 & 77.0 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{14}{c}{\rule{0pt}{10pt}\textbf{Base 150M (10B tokens)}} \\[1pt]
MLM baseline & 8192 & 90.4 & 61.3 & 94.4 & 55.5 & 79.3 & 78.9 & 88.3 & 74.9 & 79.2 & 78.6 & 68.9 & 77.3 \\
CLM detour & 8192 & 90.3 & 61.0 & 94.3 & 55.7 & 82.6 & 79.9 & 89.1 & 74.2 & 80.4 & \underline{79.3} & 69.4 & 77.8 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{14}{c}{\rule{0pt}{10pt}\textbf{Base 150M (50B tokens)}} \\[1pt]
MLM baseline & 8192 & \bfseries 90.5 & \underline{61.6} & \underline{95.1} & 55.7 & 81.6 & 80.0 & 88.7 & 74.9 & 80.3 & 78.7 & 67.1 & 77.7 \\
CLM detour & 8192 & 90.1 & \bfseries 61.9 & \bfseries 95.2 & 54.2 & \underline{83.2} & 81.0 & 89.1 & 74.5 & 80.1 & 78.8 & \underline{70.0} & \underline{78.0} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{14}{c}{\rule{0pt}{10pt}\textbf{Large 350M (50B tokens)}} \\[1pt]
MLM baseline & 8192 & \bfseries 90.5 & 61.0 & 94.9 & 55.0 & 82.3 & 82.0 & 89.4 & \bfseries 75.5 & \underline{81.8} & 76.4 & 67.8 & 77.9 \\
CLM detour & 8192 & 90.4 & 61.3 & 94.7 & \bfseries 56.5 & \bfseries 84.2 & \underline{83.2} & \bfseries 89.8 & \underline{75.3} & 81.7 & \bfseries 79.7 & 69.3 & \bfseries 78.7 \\
\bottomrule
\end{tabular}
}
\end{table*}

\Cref{tab:english-results} shows the English results at three scales. CLM outperforms MLM on average at all scales, with the gap widening at Large scale (+0.8pp, 7/11 task wins). The smaller English effect is expected: ModernBERT was pretrained on web data that commonly includes biomedical corpora such as PubMed, so its low layers already partially encode biomedical features. ModernCamemBERT, by contrast, was pretrained on general French web text without biomedical sources, leaving a larger domain gap and correspondingly stronger CLM benefit (+2.9pp vs +0.3pp). We release the English models as ModernBERT-bio.

\subsection{Optimal Decay Ratio}

We sweep the decay length from 2.5\% to 50\% of the CLM phase at two scales, evaluating on 3 French tasks (DiaMED, FrACCO-30, FrACCO-100; \Cref{tab:decay-sweep}). At both scales, 10\% decay gives the best performance; shorter decay (2.5--4\%) scores about 1.0pp below optimal, and longer decay (20--50\%) provides no additional benefit. This ratio matches the cooldown lengths used in language model pretraining.

\begin{table}[t]
\centering
\small
\caption{\textbf{Decay ratio sweep} on 3 French tasks (DiaMED, FrACCO-30, FrACCO-100; 9 seeds each). 10\% decay is optimal at both scales.}
\label{tab:decay-sweep}
\begin{tabular}{lccc}
\toprule
\textbf{Decay tokens} & \textbf{Decay / Phase~1} & \textbf{CLM+decay $F_1$} & \textbf{MLM+decay $F_1$} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{10B Phase~1 (Base, 150M)}} \\[1pt]
0.25B & 2.5\% & 71.6\% & 69.1\% \\
\textbf{1.0B} & \textbf{10\%} & \textbf{72.6\%} & 70.4\% \\
2.0B & 20\% & 71.9\% & 70.4\% \\
3.0B & 30\% & 72.1\% & 69.7\% \\
5.0B & 50\% & 71.7\% & 70.4\% \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{25B Phase~1 (Large, 350M)}} \\[1pt]
1.0B & 4\% & 73.2\% & 70.1\% \\
\textbf{2.5B} & \textbf{10\%} & \textbf{73.5\%} & 70.5\% \\
5.0B & 20\% & 72.7\% & 70.1\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Needle-in-a-Haystack}

% TODO: include needle figure from figures/output/fig2_needle.pdf

Long-context encoding is critical in biomedical NLP, where clinical documents routinely span thousands of tokens and key information (diagnoses, coding labels) can appear anywhere in the document. We design a synthetic needle-in-a-haystack task in French to evaluate long-context retrieval without entity-type confounds.

\paragraph{Protocol.} A medical fact (the ``needle'') is inserted into a clinical document (the ``haystack'') at a controlled position, and the model must determine whether a given query fact is present (binary classification). Negative examples use the same template with different slot values, so the task requires precise retrieval rather than shallow pattern matching. We define 8 French medical fact templates covering drugs, vital signs, lab results, diagnoses, procedures, allergies, treatment duration, and symptoms. Each template has 2--4 slots filled from small lexicons:

\begin{itemize}
  \item \textit{Le patient a re\c{c}u 200\,mg de morphine par voie intraveineuse.}\\
        \textbf{Distractor:} \textit{Le patient a re\c{c}u 500\,mg de tramadol par voie orale.}
  \item \textit{La tension art\'erielle mesur\'ee \'etait de 160/90\,mmHg.}
  \item \textit{Le diagnostic retenu est cirrhose h\'epatique de stade~III.}
  \item \textit{Le patient rapporte une dyspn\'ee \'evoluant depuis une semaine.}
\end{itemize}

We generate 1,500 balanced positive/negative pairs across 5 lengths (512--8,192 tokens) and 3 positions (start, middle, end), split 70/15/15 for train/validation/test. We freeze each encoder and train a 2-layer MLP probe (Dropout $\to$ Linear $\to$ GELU $\to$ Dropout $\to$ Linear) on CLS representations for 3 epochs (lr $= 2 \times 10^{-5}$, batch size 4, AdamW), selecting the best checkpoint by validation accuracy.

\paragraph{Results.} CLM outperforms MLM at every context length and position (+10.7pp overall: 62.2\% vs 51.6\%). MLM accuracy degrades monotonically with document length (57\% at 512 tokens to 43\% at 8,192), while CLM degrades more slowly and remains above MLM throughout. The CLM advantage is largest at mid-document positions, where the inserted fact is farthest from both sequence boundaries. This is consistent with the downstream pattern: the French tasks with the largest CLM gains (FrACCO, CANTEMIST) use sequence lengths of 4,096 tokens during fine-tuning.


\section{Analysis: Why Does the CLM Detour Work?}
\label{sec:analysis}

\subsection{CKA Methodology}
\label{sec:cka_method}

We measure representational similarity with linear Centered Kernel Alignment~\citep{kornblith2019similarity}. Given two representation matrices $X \in \mathbb{R}^{n \times p}$ and $Y \in \mathbb{R}^{n \times q}$ (one per model, $n$ samples, mean-pooled over tokens), we center each column, compute Gram matrices $K = XX^\top$ and $L = YY^\top$, and define:
\[
\text{CKA}(X, Y) = \frac{\text{tr}(KL)}{\sqrt{\text{tr}(K^2)\,\text{tr}(L^2)}}
\]
We report divergence $d = 1 - \text{CKA}$, so higher values indicate greater representational difference. CKA equals 1 when representations are identical up to invertible linear transformation, 0 when there is no linear relationship. All computations use float64 arithmetic. For French we use 500 held-out texts from DiaMED and FrACCO; for English, PubMed abstracts. Results are averaged over 3 seeds.

\subsection{The CLM Imprint Persists in Low Layers}
\label{sec:persistence}

% TODO: include CKA figure from figures/output/fig10_cka_combined.pdf

Comparing layer-by-layer CKA between CLM-detour and MLM-only models, the CLM phase modifies low-layer representations in a way that subsequent MLM training does not undo, even when the decay budget matches the CLM phase. CKA divergence reaches approximately 56.5\% after 1.5B tokens of MLM decay and remains stable, with only 0.6pp variation over 8B additional tokens.

To isolate where CLM differs from MLM \emph{specifically} rather than just from training noise, we normalize each layer's CLM-MLM divergence by a seed-noise control (two MLM models trained with different random seeds on identical data):
\[
r_\ell = \frac{1 - \text{CKA}(\text{CLM}_\ell,\; \text{MLM}^{s_1}_\ell)}{1 - \text{CKA}(\text{MLM}^{s_2}_\ell,\; \text{MLM}^{s_1}_\ell)}
\]
A ratio of 1 means no CLM-specific effect. Low layers (0--7) show ratios of 5--44$\times$: CLM changes them far more than random seed variation does. Mid and deep layers are near 1$\times$: both objectives modify them similarly, so CKA cannot distinguish CLM from MLM there.

\subsection{Causal Evidence: Freeze Interventions}
\label{sec:freeze-results}

% TODO: include freeze figure from figures/tikz_figure1.tex

Do these low-layer changes actually drive the downstream improvement? We answer with the freeze interventions described in \S\ref{sec:freeze}, summarized in \Cref{tab:freeze}.

\begin{table}[t]
\centering
\small
\caption{\textbf{Freeze intervention results} (French Base, 8 tasks, 9 seeds). Freezing low layers during CLM eliminates the benefit ($p=0.36$ vs MLM baseline). Freezing mid layers preserves it. Freezing low layers during decay has negligible effect.}
\label{tab:freeze}
\begin{tabular}{lcc}
\toprule
\textbf{Condition} & \textbf{Avg $F_1$} & $\boldsymbol{\Delta}$ \textbf{vs MLM} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{3}{c}{\rule{0pt}{9pt}\textbf{References}} \\[1pt]
MLM baseline & 57.9 & --- \\
CLM detour & \bfseries 60.8 & +2.9 \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{3}{c}{\rule{0pt}{9pt}\textbf{Freeze interventions}} \\[1pt]
Freeze L0--7 (CLM phase) & 58.4 & +0.5 \\
Freeze L8--14 (CLM phase) & 60.3 & +2.4 \\
Freeze L0--7 (decay phase) & 60.3 & +2.4 \\
\bottomrule
\end{tabular}
\end{table}

Freezing low layers (0--7) during CLM drops $F_1$ from 60.8\% to 58.4\%. We cannot reject the null hypothesis that the resulting model matches the MLM baseline ($p = 0.36$, paired bootstrap across 8 tasks $\times$ 9 seeds). Freezing mid layers (8--14) during CLM preserves the full benefit (60.3\%). Together, these two experiments establish a double dissociation: low layers are necessary and mid layers are not. Freezing low layers during decay has negligible effect ($-$0.5pp, 60.3\%), confirming that MLM decay preserves the CLM imprint in low layers.

\subsection{Non-Localizability}

The freeze experiments show that low layers are necessary \emph{during training}. We now ask whether the resulting benefit can be localized \emph{after training}, by copying full parameter blocks from the CLM model into the MLM model.

\begin{table}[t]
\centering
\small
\caption{\textbf{Layer transplantation} on DiaMED (3 seeds, linear probing). No transplant recovers more than 35\% of the 7.1pp gap.}
\label{tab:transplant}
\begin{tabular}{lccc}
\toprule
\textbf{Transplant} & \textbf{$F_1$ (\%)} & $\boldsymbol{\Delta}$ & \textbf{Recovery} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{References}} \\[1pt]
MLM baseline & 32.8 & --- & 0\% \\
CLM detour & 39.9 & +7.1 & 100\% \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{Layer group transplants}} \\[1pt]
Low (0--7) & 31.9 & $-$0.9 & $-$12\% \\
Mid (8--14) & 35.3 & +2.5 & +35\% \\
Late (15--21) & 32.4 & $-$0.4 & $-$6\% \\
\bottomrule
\end{tabular}
\end{table}

No transplant recovers more than 35\% of the gap (\Cref{tab:transplant}). Low and late layer transplants hurt performance. Inserting CLM low layers into an MLM model creates a mismatch with the mid and deep layers that co-adapted under MLM, yielding representations worse than either pure model.

Component-level transplants (\Cref{tab:transplant-comp}) confirm the same pattern. Replacing all attention modules from CLM into MLM recovers only 15\% of the gap; replacing all MLP blocks recovers 35\%. No single component type captures the full benefit.

\begin{table}[t]
\centering
\small
\caption{\textbf{Component-level transplants} on DiaMED (3 seeds, linear probing). Neither attention nor MLP modules alone recover the full gap.}
\label{tab:transplant-comp}
\begin{tabular}{lccc}
\toprule
\textbf{Transplant} & \textbf{$F_1$ (\%)} & $\boldsymbol{\Delta}$ & \textbf{Recovery} \\
\midrule
All attention & 33.9 & +1.1 & +15\% \\
All MLP & 35.3 & +2.5 & +35\% \\
\bottomrule
\end{tabular}
\end{table}

Unlike freeze interventions, which act during training, transplants replace already-trained weights and cannot reconstruct the cross-layer co-adaptation that developed during CLM. The freeze experiments show low layers are necessary during training; the transplant experiments show the resulting benefit cannot be isolated post hoc.

\subsection{Scaling and Architectural Asymmetry}
\label{sec:asymmetry}

\begin{table}[t]
\centering
\small
\caption{\textbf{CKA divergence scales with model size and is asymmetric.} The encoder-decoder comparison uses different model families (see text).}
\label{tab:asymmetry}
\begin{tabular}{llcc}
\toprule
\textbf{Model} & \textbf{Params} & \textbf{Div.} & \textbf{CI} \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{Encoders (MLM$\to$CLM$\to$MLM)}} \\[1pt]
ModernCamemBERT Base & 150M & 56.5\% & $\pm$0.7\% \\
ModernCamemBERT Large & 350M & 67.2\% & $\pm$0.3\% \\
\midrule
\rowcolor{ThesisTableSep} \multicolumn{4}{c}{\rule{0pt}{9pt}\textbf{Decoder (CLM$\to$MLM$\to$CLM)}} \\[1pt]
Gemma-3 & 270M & 26.3\% & $\pm$1.0\% \\
\bottomrule
\end{tabular}
\end{table}

The CLM imprint increases with model capacity (\Cref{tab:asymmetry}): Large (350M) reaches 67.2\% $\pm$ 0.3\% divergence versus 56.5\% $\pm$ 0.7\% for Base (150M). Testing the reverse on Gemma-3 (270M decoder), MLM-then-CLM retains only 26.3\% $\pm$ 1.0\% divergence versus 67.2\% for an encoder of similar scale. One explanation is the density of the decay signal: MLM decay at 15\% masking provides weak erasure, while CLM decay at 100\% provides strong erasure. This comparison is exploratory: it is confounded by model family and language, and a controlled comparison using encoder and decoder variants of the same family would be needed to confirm the asymmetry.


\section{Limitations}

Our freeze interventions use French Base models only. The CKA patterns that motivate them (low-layer divergence well above seed noise) are consistent across French Base, French Large, and English Base, but replication of the freeze interventions on English and Large models would strengthen the causal claim. The encoder-decoder asymmetry comparison is confounded by model family. The position trade-off at 2048+ tokens means that the approach may not be suitable for tasks requiring very long context, such as full discharge summary coding.


\section*{Conclusion}

Training an encoder like a decoder, then converting it back, produces stronger representations than training with MLM alone. The CLM detour reshapes low transformer layers in ways that subsequent MLM decay does not undo: a lasting imprint that scales with model capacity and produces consistent downstream gains across 8 French and 11 English biomedical tasks.

The practical implication is straightforward: for biomedical encoder pretraining, the CLM detour is a free lunch. Same data, same compute, better results. MLM decay of roughly 10\% of the detour length suffices, and the effect is strongest when the domain gap between pretraining and target data is large.

\vspace{2em}

The CLM detour improves the encoder by changing the pretraining objective. Another lever, complementary to this one, is the pretraining data itself. So far, biomedical adaptation has consumed unstructured text: articles, leaflets, clinical cases. Yet medicine also encodes a substantial body of structured knowledge in clinical ontologies such as ICD-10, SNOMED, and ATC. This knowledge enters the model only indirectly, through whatever scientific prose happens to mention it. In the next chapter, we ask whether this structured knowledge can be injected directly during pretraining.
