\begin{center}
  \textsc{Abstract}
\end{center}
\noindent

Hospitals accumulate large volumes of clinical documents that record diagnoses, treatments, test results, and the reasoning that connects them. Converting their contents into structured variables would support clinical research, but these documents are highly sensitive, rarely shareable, and therefore almost absent from publicly available French-language resources. This thesis investigates how compact information-extraction models can be trained without hospital records and designed for local use within the institutions where those records remain.

We first determine which public texts can provide relevant training data. We assemble and release French biomedical corpora from public sources. We also annotate the multilingual, predominantly English PubMed Central corpus at paragraph level by content type and quality, because its scientific articles mix relevant medical content with boilerplate. On downstream tasks, pretraining on the selected paragraphs matches pretraining on the full corpus while using one third as many tokens. This annotation also identifies two million public clinical-case passages across languages, most of them in English. We finally show that generic neural quality filters can increase overlap between training data and evaluation benchmarks, raising the risk that models are tested on examples they have already seen.

We then train compact biomedical encoders. Continuing a general French model's pretraining on biomedical text improves its performance on all five benchmarks tested, without training a new model from scratch. We next introduce a causal detour: a temporary autoregressive phase in which the model learns through next-token prediction before returning to standard bidirectional masked language modelling. This strategy outperforms comparable masked-language-model training on all eight French biomedical tasks, and the same effect holds in English. We also introduce OntoBook, which complements text pretraining by converting medical ontologies into synthetic textbooks; it improves document classification and medical coding. We release the resulting models and data.

To turn these encoders into useful systems, we address the downstream task that motivates this thesis: extracting study-specific variables from clinical documents. Each project asks for different fields, usually with little or no annotated data. We use large language models to annotate public text, to evaluate extractions for entity types absent from existing annotated benchmarks, and to generate synthetic lymphoma dossiers derived from clinical cases published in scientific articles. Clinicians reviewed these dossiers through a web-based annotation tool and found them plausible overall. The generated supervision trains compact open-vocabulary extractors: their target fields are described in natural language at inference rather than fixed during training. We release the general-purpose extractors and a synthetic benchmark. On this benchmark, our released 150-million-parameter extractor outperforms extractive question answering; its $F_1$ score on extracted values comes within two points of that of the strongest generator fine-tuned on the same data, despite being twenty-seven times smaller.

Together, these results show that compact clinical extractors can be built from carefully selected public data and synthetic supervision, without treating model scale as a substitute for data design. Transfer to real hospital records remains to be fully established. Finally, the rapid evolution of foundation models and the emergence of agents that can consult records and use clinical tools make realistic evaluation of their clinical safety an urgent research priority.

\clearpage
\selectlanguage{french}
\begin{center}
  \textsc{Résumé}
\end{center}
\noindent

Les hôpitaux accumulent des documents cliniques qui consignent les diagnostics, les traitements, les résultats d'examens et le raisonnement qui les relie. Convertir leur contenu en variables structurées faciliterait la recherche clinique. Cependant, ces documents sont sensibles, rarement partageables et donc presque absents des ressources publiques en français. Cette thèse étudie comment entraîner, sans dossiers hospitaliers, des modèles compacts d'extraction d'information conçus pour une utilisation locale dans les établissements qui conservent ces dossiers.

Nous déterminons d'abord quels textes publics fournissent des données d'entraînement pertinentes et constituons plusieurs corpus biomédicaux français. Nous annotons aussi par type et qualité les paragraphes du corpus multilingue PubMed Central, majoritairement anglais, dont les articles mêlent contenu médical et texte générique. Sur les tâches évaluées, pré-entraîner le modèle sur les paragraphes sélectionnés atteint les mêmes performances qu'utiliser le corpus complet, avec un tiers des tokens. Cette annotation identifie aussi deux millions de passages de cas cliniques dans plusieurs langues, en majorité en anglais. Nous montrons enfin que des filtres neuronaux génériques de qualité peuvent accroître le chevauchement entre données d'entraînement et jeux d'évaluation, au risque d'évaluer les modèles sur des exemples déjà vus.

Nous entraînons ensuite des encodeurs biomédicaux compacts. Poursuivre le pré-entraînement d'un modèle généraliste français sur des textes biomédicaux améliore ses performances sur les cinq jeux d'évaluation étudiés, sans repartir de zéro. Nous introduisons ensuite un détour causal : une phase autorégressive temporaire, où le modèle apprend par prédiction du prochain token, avant de revenir au masquage bidirectionnel standard. Cette stratégie surpasse un entraînement comparable par masquage sur les huit tâches biomédicales en français ; le même effet est observé en anglais. OntoBook complète enfin le pré-entraînement textuel en transformant des ontologies médicales en manuels synthétiques, ce qui améliore la classification de documents et le codage médical. Nous publions les modèles et les données obtenus.

Nous appliquons ces encodeurs à la tâche qui motive cette thèse : extraire de documents cliniques les variables propres à une étude. Les champs varient selon les projets et les données annotées sont rares ou absentes. Nous utilisons de grands modèles de langue pour annoter des textes publics, évaluer les extractions pour des types d'entités absents des jeux annotés existants et générer des dossiers synthétiques de lymphome à partir de cas cliniques publiés. Des cliniciens ont examiné ces dossiers avec un outil en ligne et les ont jugés globalement plausibles. Cette supervision entraîne des extracteurs compacts à vocabulaire ouvert : les champs recherchés sont décrits en langage naturel au moment de l'utilisation, plutôt que fixés lors de l'entraînement. Nous publions ces extracteurs et un jeu d'évaluation synthétique. Sur ce jeu d'évaluation, notre extracteur publié de 150 millions de paramètres surpasse les approches par questions-réponses extractives ; sa F-mesure sur les valeurs extraites reste à moins de deux points de celle du meilleur modèle génératif affiné sur les mêmes données, qui compte 27 fois plus de paramètres.

Ces résultats montrent que des données publiques sélectionnées et une supervision synthétique permettent de construire des extracteurs compacts pour des tâches cliniques, sans remplacer le choix des données par une augmentation de la taille du modèle. Leur transfert aux dossiers hospitaliers réels reste à établir. Enfin, l'évolution rapide des modèles de fondation et l'apparition d'agents capables de consulter des dossiers et d'utiliser des outils cliniques rendent urgente une évaluation réaliste de leur sécurité clinique.
\selectlanguage{english}
