%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Modern AI and New Opportunities}
\label{chap:modern-ai}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% =====================================================================
% TODO — chapter to be written (scribe pass, voice-first). Scaffold only.
% Arc: where contemporary AI now stands -> what it makes newly possible
%      in the clinic (OncoLab) -> the wall it hits (no shareable data),
%      which motivates the "What is Biomedical Text?" chapter.
% =====================================================================

\paragraph{Where AI stands today.}
\begin{itemize}
  \item TODO: the shift from hand-built expert systems (MYCIN, ch.~\ref{chap:ai-health})
        to learned models; foundation models / LLMs as general-purpose engines.
  \item TODO: why this matters for medicine now — fluent language, in-context
        adaptation, broad coverage out of the box.
\end{itemize}

\paragraph{A concrete opportunity: the OncoLab project.}
\begin{itemize}
  \item TODO: describe OncoLab — goal, clinical setting, the information-extraction
        task (oncology reports / structuring).
  \item TODO: what modern AI makes newly feasible here that was out of reach for
        rule-based systems.
  \item TODO: the practical stakes (cohort selection, report structuring, decision
        support).
\end{itemize}

\paragraph{The challenge: opportunity without data.}
\begin{itemize}
  \item TODO: the wall — clinical text is personal data, not shareable; no large
        real French clinical corpus to train or even evaluate on.
  \item TODO: so the most direct lever (just fine-tune on in-domain data) is
        unavailable; we must build from public text instead.
  \item TODO: this reframes the question — before asking how to model biomedical
        text, we must ask what biomedical text actually \emph{is}, and what public
        sources can stand in for the clinic.
\end{itemize}

% Transition into the triptych (ch. "What is Biomedical Text?").
