%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\partepigraph{``The biggest lesson that can be read from 70 years of AI research is
that general methods that leverage computation are ultimately the most effective, and
by a large margin. [\ldots] We should build in only the meta-methods that can find and
capture this arbitrary complexity. [\ldots]''}{Richard S. Sutton, \textit{The Bitter Lesson}}

This thesis asks how to train language models for clinical text when direct
clinical data is scarce. The opening part sets up that problem. The first
chapter traces how healthcare AI moved from hand-built medical knowledge to
learning from large collections of text, and why the clinical setting resists
that shift. The next chapter looks at what current systems make possible and
at the data shortage they run into. A short final chapter states the
contributions of the thesis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Artificial Intelligence and Healthcare}
\label{chap:ai-health}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent
\begin{center}
\begin{tikzpicture}
\node[draw=ThesisNeutral, fill=ThesisPaper, rounded corners=3pt,
      line width=0.5pt, inner sep=12pt]{%
\begin{minipage}{0.82\textwidth}
\footnotesize\color{ThesisInk}
If a physician is asked, ``How do you make a medical diagnosis?'' his explanation
of the process might be as follows. ``First, I obtain the case facts from the
patient's history, physical examination, and laboratory tests. Second, I evaluate
the relative importance of the different signs and symptoms. [\ldots] Third, to
make a differential diagnosis I list all the diseases which the specific case can
reasonably resemble. Then I exclude one disease after another from the list until
it becomes apparent that the case can be fitted into a definite disease category
[\ldots]''
\end{minipage}%
};
\end{tikzpicture}
\captionof{figure}{A physician's own account of how a diagnosis is made~\citep{ledley_lusted_1959}.}
\label{fig:ledley-lusted}
\end{center}
\medskip

The quotation comes from Robert Ledley and Lee Lusted, writing in 1959. In \emph{Science}, they argued that diagnosis could be made explicit enough to be written down and, in principle, carried out by a machine~\citep{ledley_lusted_1959}. The 1956 Dartmouth workshop had named ``artificial intelligence'' only three years before, with broad goals such as language and learning; medicine sat outside that initial agenda. Ledley and Lusted framed diagnosis as a medical version of the same problem: a form of reasoning that could be specified and run by a machine. A few years later, programs began to implement this idea.

In 1961, Homer Warner and his colleagues built a system that used a patient's signs to estimate the probability of congenital heart disease, one of the first working versions of the new approach~\citep{warner_1961}. A decade later, F.~T. de Dombal's group in Leeds tested a program on the acute abdomen, the urgent problem of distinguishing appendicitis from other causes of abdominal pain. They reported that it matched the attending clinicians and, in their trial, slightly exceeded them~\citep{dedombal_1972}. The performance depended on the data the program had seen: the probabilities were only as good as the cases used to estimate them.

At Stanford in the 1970s, MYCIN advised on antibiotics for blood infections from a set of hand-written rules. In a blinded evaluation, experts rated its recommendations acceptable in 65\% of cases, against 42.5 to 62.5\% for the human prescribers~\citep{mycin_1975, mycin_eval_1979}. INTERNIST-1, built at Pittsburgh, aimed at the whole of internal medicine and encoded hundreds of diseases. Yet its own authors, testing it on real cases, judged it ``not sufficiently reliable for clinical applications''~\citep{internist1_1982}.

These systems had two main limits. Their knowledge was written by hand: every rule had to be drawn from experts and kept up to date as medicine changed. This made the programs expensive to build and brittle outside the narrow cases they were designed for. They also remained hard to use in the clinic. Physicians wanted reasoning they could check; opaque verdicts from an oracle were a poor fit~\citep{teach_shortliffe_1981}. There was as yet no electronic medical record for such a program to draw on. Public funding for artificial intelligence collapsed twice in these decades, and medical expert systems left routine care largely untouched.

Machine learning changed the source of knowledge. Systems began to learn directly from data instead of encoding what experts knew, and the gains were large. By 2016, a neural network read retinal photographs for diabetic retinopathy about as accurately as ophthalmologists~\citep{gulshan_2016}, and a related approach matched dermatologists at spotting skin cancer from images~\citep{esteva_2017}. The same shift reached language, where systems trained on large amounts of text became the best way to process it. Medicine had found that the strongest approach set aside much of the hand-built knowledge the field had spent decades assembling.
