\section{Introduction}

The previous chapter produced training data without a fixed label set, and the introduction to this part explained why a fixed label set does not work for clinical extraction: the types we want change from one study to the next, and most will never have been seen in training. A model with a fixed output layer cannot produce a type it was not trained on. This chapter builds a model with no fixed output layer. Instead of choosing among a closed list of labels, it compares a span of text against a short description of the type we are looking for and decides whether they match. The type is given at the moment of use, so one model can extract a disease, a dosage, or a treatment response without being retrained for each.

\section{Open-Vocabulary Extraction by Type Description}

This is the idea behind GLiNER~\citep{zaratiana2024gliner}, which works by comparison. It reads the document together with the descriptions of the types we are looking for, and scores how well each span of text matches each type. There is no fixed output layer with one slot per label. Because the types are supplied as text at inference time, the set of types is open: a user can ask for a type the model never saw in training, and the model will still try to find it.

Our extractor, \emph{ModernCamemBERT-bio-gliner-base}, follows the GLiNER2 design~\citep{zaratiana2025gliner2}. A single encoder reads the document and the candidate type descriptions in one pass, and a light span-scoring head matches every span against every type. The same machinery also handles the other tasks the training data carries: classifying the document, filling structured fields, and extracting relations between spans.

We release the model in two configurations that differ only in their training-data mix. The first is tuned for entity extraction and is the stronger general extractor: it finds types it was never trained on more accurately than any baseline we compare against in \Cref{chap:evaluation}. The second is multi-task: alongside entities it classifies the document, fills structures, and extracts relations, and it does these better than the first, at some cost to entity extraction. This trade-off between the two is one of the findings of this part, and we return to it in \Cref{chap:evaluation}.

\section{Recognition as Similarity Matching}

The matching is only as good as the encoder that places type descriptions and spans in a shared space. GLiNER turns recognition into a comparison: a label encoder reads a type description and produces a vector, a document encoder reads a candidate span and produces a vector, and a cosine similarity decides whether the two refer to the same thing. \Cref{fig:medembed-schema} shows the two paths. Because the type is supplied as text, a description the model never saw in training still produces a vector, and the same span can be tested against any number of types at once. Everything depends on whether the two vectors are close when the type matches the span and far when it does not. The embedding space sets the limit, so we build an encoder for it.

\begin{center}
\begin{tikzpicture}[
    font=\small\color{ThesisInk},
    box/.style={draw=ThesisNeutral, rounded corners=3pt, line width=0.6pt,
                inner sep=7pt, align=center, minimum height=11mm},
    enc/.style={box, fill=ThesisPrimary!18, draw=ThesisPrimaryDark, minimum width=26mm},
    txt/.style={box, fill=ThesisPaper, minimum width=34mm},
    vec/.style={box, fill=ThesisSecondary!18, draw=ThesisSecondaryDark, minimum width=20mm},
    flow/.style={->, line width=1pt, ThesisInk!45, >=stealth, shorten >=2pt, shorten <=2pt}
]
% top path: label
\node[txt] (lab) at (0,1.7) {\textit{Disease: a pathological}\\\textit{condition affecting an organ}};
\node[enc] (le) at (4.6,1.7) {Label encoder};
\node[vec] (lv) at (8.2,1.7) {vector};
% bottom path: span
\node[txt] (sp) at (0,-1.7) {\textit{infarctus du myocarde}\\(span in the document)};
\node[enc] (de) at (4.6,-1.7) {Document encoder};
\node[vec] (sv) at (8.2,-1.7) {vector};
% arrows
\draw[flow] (lab) -- (le);
\draw[flow] (le) -- (lv);
\draw[flow] (sp) -- (de);
\draw[flow] (de) -- (sv);
% match
\node[box, fill=ThesisTertiary!22, draw=ThesisTertiaryDark, minimum width=24mm] (cos) at (11.4,0) {cosine\\similarity};
\draw[flow] (lv) -- (cos);
\draw[flow] (sv) -- (cos);
\node[align=center, ThesisNeutral, font=\footnotesize] at (11.4,-1.7) {high if the type\\matches the span};
\end{tikzpicture}
\captionof{figure}{Recognition as similarity matching: a type description and a text span are each mapped to a vector, and their cosine similarity decides the match.}
\label{fig:medembed-schema}
\end{center}

\section{What a Good Label Encoder Must Do}

The label encoder does most of the work. It has to read a medical type description and place it near the spans that belong to that type, even for types it never saw in training, and it has to do this in French, where biomedical text is scarce. The encoder must understand what a medical term means instead of matching surface words, so that a code, its label, and a clinical synonym land in the same region. It must also keep close terms apart when they are clinically distinct, since many medical terms differ only in small ways.

A plainly pretrained French encoder fails this test. On the terminology-coding tasks we describe below, a general ModernCamemBERT scores barely above chance: it knows French, but it has not learned which terms denote the same medical concept and which only look alike. MedEmbed targets this grounding of French biomedical terminology, which neither a general French encoder nor a strong multilingual one covers well.

\section{Training Data}

MedEmbed is trained on description-span pairs assembled from four families, summarised in \Cref{tab:medembed-data}. The first family, terminology pairs, is built from the French nomenclatures CIM-10, CCAM, and ATC by the ontology extraction of \Cref{chap:ontobook}; we describe the kinds of pairs here but do not repeat the extraction. From the structured graph of each terminology we read out, for every code, its official label, its synonyms, its parent and child codes, and the codes it is explicitly distinguished from. A code paired with its label or a synonym teaches the encoder that they denote the same concept. A child paired with its parent teaches the hierarchy. An exclusion link, where the terminology states that one code must not be confused with a named other, gives a hard negative: a positive and a deliberately close wrong answer for the same anchor.

The other three families come from \Cref{chap:synthetic}. Synthetic passages are short question-and-answer pairs generated from French biomedical articles, where the question is the anchor and a relevant passage is the positive. Long documents are the same idea at full length, with the positive being a whole document, which exercises the 8192-token context. Persona queries take one source document and ask about it from several points of view, so that a clinician's technical question and a patient's plain question both map to the same passage. \Cref{fig:medembed-examples} shows real pairs from each family.

\begin{table}[t]
\centering
\caption{The four families of training pairs and their sizes.}
\label{tab:medembed-data}
\sisetup{group-separator={,}, group-minimum-digits=4}
\begin{tabular}{l S[table-format=6.0]}
\toprule
Source & {Pairs} \\
\midrule
Terminology pairs (CIM-10, CCAM, ATC) & 258521 \\
Synthetic passages & 489680 \\
Persona queries & 187310 \\
Long documents & 25638 \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}
\usetikzlibrary{matrix}
% Each example pair gets its own panel: grey sffamily tag, then the
% anchor/positive(/negative) lines, with whitespace between panels.
\newcommand{\extag}[1]{{\scriptsize\sffamily\color{ThesisNeutral} #1}\\[3pt]}
\newcommand{\exrole}[1]{{\scriptsize\sffamily\color{ThesisNeutral} #1}}
\newcommand{\expanel}[2]{%
  \begin{minipage}{\dimexpr\textwidth-22pt\relax}
  \footnotesize\color{ThesisInk}
  \extag{#1}#2
  \end{minipage}}
\begin{tikzpicture}[
    panel/.style={draw=ThesisNeutral!55, fill=ThesisPaper, rounded corners=2.5pt,
                  line width=0.5pt, inner sep=8pt}]
\matrix[matrix of nodes, nodes={anchor=west}, row sep=5pt, column sep=0pt,
        every node/.style={panel}] {
  \expanel{CODE $\rightarrow$ LABEL \textnormal{(CIM-10)}}{%
    \exrole{anchor} \textit{B08.5}\\[1pt]
    \exrole{positive} \textit{Pharyngite vésiculaire due à un entérovirus}} \\
  \expanel{SYNONYM \textnormal{(CIM-10)}}{%
    \exrole{anchor} \textit{B08.5 Pharyngite vésiculaire due à un entérovirus}\\[1pt]
    \exrole{positive} \textit{Herpangine}} \\
  \expanel{HIERARCHY \textnormal{(child $\rightarrow$ parent, CIM-10)}}{%
    \exrole{child} \textit{F02.00 Démence de la maladie de Pick, sans symptôme supplémentaire}\\[1pt]
    \exrole{parent} \textit{F02.0 Démence de la maladie de Pick}} \\
  \expanel{EXCLUSION \textnormal{(hard negative: anchor / positive / negative, CIM-10)}}{%
    \exrole{anchor\hphantom{x}} \textit{D37.7 Tumeur à évolution imprévisible ou inconnue d'autres organes digestifs}\\[1pt]
    \exrole{positive} \textit{Tumeur à évolution imprévisible ou inconnue d'autres organes digestifs}\\[1pt]
    \exrole{negative} \textit{D48.5 Tumeur à évolution imprévisible et inconnue de la peau}} \\
  \expanel{PERSONA \textnormal{(clinician vs patient, same source document)}}{%
    \exrole{clinician} \textit{Quelle est la corrélation entre le taux des IgE sériques totales et la gravité de la dermatite atopique ?}\\[1pt]
    \exrole{patient\hphantom{xx}} \textit{Est-ce que l'érythème noueux est grave docteur ?}} \\
};
\end{tikzpicture}
\captionof{figure}{Real training pairs: a code and its label, a synonym, a child-to-parent hierarchy link, an exclusion triplet with a deliberately close wrong answer, and the same source document queried by a clinician and by a patient.}
\label{fig:medembed-examples}
\end{center}

\section{Contrastive Training}

The encoder is trained to pull a matching pair together and push the rest apart. Most of the data is simple pairs with no curated negative, and for these we use a multiple-negatives ranking loss. For each anchor in a batch, its own positive is the right answer and every other positive in the same batch is treated as a wrong answer, so a large batch supplies many negatives for free and the contrastive signal grows with batch size. This suits terminology, where most pairs are a positive match and the hard cases are rare.

The rare cases are handled by a triplet loss on the curated hard negatives: the exclusion links from the terminologies, sibling reformulations under the same parent, and negatives mined by searching for the most similar wrong passage. The random in-batch negatives of the first loss are usually easy, because two terminology entries picked at random are rarely confusable. The triplet loss supplies the difficulty the data was built to teach, forcing a gap between a positive and a known-wrong-but-similar negative, with a cosine distance that matches the cosine used at inference.

The released model is trained in two stages, terminology pairs first. Adding the passages and documents in a second stage does not improve terminology coding and slightly hurts it, so the second stage is dropped and the terminology-only model is released. This shows that the terminology pairs carry the signal that matters. The backbone is ModernCamemBERT-bio from \Cref{chap:objectives,chap:ontobook}, with mean pooling over tokens followed by L2 normalisation, 149 million parameters, and an 8192-token context.

\section{Results}

\Cref{tab:medembed-results} compares MedEmbed against three strong baselines on OntoBench-FR, a retrieval benchmark built from the French terminologies and from real clinical cases. MedEmbed reaches the highest weighted average and beats encoders several times its size, including the multilingual mE5-large-instruct and BGE-M3 and the strongest general French encoder, Solon. The results give two findings. First, contrastive training matters more than size: a 149-million-parameter encoder trained this way beats a raw billion-parameter model that was never trained to compare, so the way the space is shaped matters more than the number of parameters. Second, the starting encoder matters, since ModernCamemBERT-bio gives a clear gain over a plainly pretrained encoder on the novel types that generalisation depends on.

\begin{table}[t]
\centering
\caption{OntoBench-FR accuracy, weighted average over benchmarks.}
\label{tab:medembed-results}
\sisetup{detect-weight=true, mode=text}
\begin{tabular}{l S[table-format=2.1]}
\toprule
Model & {Avg.} \\
\midrule
\tablegroup{2}{Our model}
MedEmbed (149M) & \bfseries 58.1 \\
\tablegroup{2}{Baselines}
mE5-large-instruct & 48.9 \\
BGE-M3 & 47.2 \\
Solon & 44.9 \\
\bottomrule
\end{tabular}

\vspace{2pt}
{\footnotesize\color{ThesisNeutral} Average accuracy over the four benchmarks, weighted by benchmark size.}
\end{table}

\Cref{tab:medembed-ablation} isolates each data family by training on it alone or removing it. The numbers here are an unweighted grand average over the four benchmarks, on a different scale from \Cref{tab:medembed-results}, so the two tables must not be compared directly. Terminology pairs are the one source that matters. Training on them alone gives the best score of the whole study, above the full mixture, while removing them costs about fifteen points and collapses the CCAM and CIM-10 coding tasks in particular. Personas, passages, and documents on their own give the weakest scores. This is why the released model is the terminology-only one: the richer data does not help, and the structured terminology pairs give a label encoder the signal it needs to tell close terms apart.

\begin{table}[t]
\centering
\caption{Data-source ablation: grand average over the four benchmarks.}
\label{tab:medembed-ablation}
\sisetup{detect-weight=true, mode=text}
\begin{tabular}{l S[table-format=2.2]}
\toprule
Training data & {Grand avg.} \\
\midrule
Terminology pairs only & \bfseries 60.31 \\
Full mixture & 58.06 \\
No personas & 56.90 \\
Personas only & 49.01 \\
\textit{Solon (baseline)} & \textit{47.89} \\
No terminology pairs & 45.60 \\
Long documents only & 45.27 \\
Synthetic passages only & 44.18 \\
\bottomrule
\end{tabular}

\vspace{2pt}
{\footnotesize\color{ThesisNeutral} Unweighted average over the four benchmarks.}
\end{table}

\section{Training the Extractor}

We train the extractor on the synthetic annotations of \Cref{chap:synthetic}, with no real clinical labels at any point. Both configurations use the same recipe and run in a few hours on a single GPU. The result is a small open model that extracts types defined at inference time, trained entirely on public and synthetic data.

\section*{Conclusion}

A model that can extract a type it was never trained on is hard to measure. The usual way to score extraction is to compare against a fixed human annotation, but a fixed annotation only covers the types the annotators were asked to mark, which is the closed world this model was built to leave behind. Measuring it fairly needs a different kind of evaluation, and that is the subject of the next chapter.
