\chapter{Lay and Web Text} \label{chap:bt-web}

\noindent
\begin{center}
% Representative (fictional, de-identified) patient-forum message; lay/involved register.
\begin{tikzpicture}
\node[draw=ThesisNeutral, fill=ThesisPaper, rounded corners=3pt, line width=0.5pt,
      inner sep=10pt] {%
\begin{tabular}{@{}l !{\hspace{6pt}\textcolor{ThesisNeutral}{\vrule width 0.3pt}\hspace{6pt}} l@{}}
\begin{minipage}[t]{0.40\textwidth}
\footnotesize\color{ThesisInk}
{\tiny\sffamily\color{ThesisNeutral} FRANÇAIS}\\[5pt]
\textit{Bonjour, mon père a fait une crise cardiaque la semaine dernière. Il a eu
une grosse douleur dans la poitrine qui descendait dans le bras gauche, il était
tout pâle et en sueur, on a appelé le 15 tout de suite. Est-ce que quelqu'un sait
combien de temps dure la convalescence après un infarctus ? Merci d'avance.}
\end{minipage}
&
\begin{minipage}[t]{0.40\textwidth}
\footnotesize\color{ThesisNeutral}
{\tiny\sffamily\color{ThesisNeutral} ENGLISH · TRANSLATION}\\[5pt]
\textit{Hello, my father had a heart attack last week. He had a strong pain in the
chest that went down into the left arm, he was all pale and sweating, we called
emergency services right away. Does anyone know how long recovery takes after a
heart attack? Thanks in advance.}
\end{minipage}
\end{tabular}%
};
\end{tikzpicture}
\captionof{figure}{A patient forum about myocardial infarction; French original, English translation.}
\label{fig:bt-web}
\end{center}
\medskip

% === P1 — HOOK ===
\begin{reviewread}
\Cref{fig:bt-web} shows the same heart attack once more, but this time in the words of the people who live through it.
There is no technical vocabulary anymore.
A worried relative writes \emph{crise cardiaque} (\emph{heart attack}) rather than \emph{infarctus du myocarde} (\emph{myocardial infarction}).
The symptoms are described in plain, everyday words.
For at least two decades the internet has made this kind of text widely available.
By 2013, 72\% of US internet users had looked online for health information \citep{fox_health_online_2013}, creating a vast and freely accessible body of patient-facing medical text.
This text is far from the language of the clinical domain, but it offers something research articles do not: an abundant and more varied representation of medical knowledge.
It is therefore worth examining how medical language, and especially medical terms, change when they reach the public. 
\end{reviewread}

% === P2 — Fleck ===
\begin{reviewread}
The idea that a fact changes as it travels is an old one.
In 1935 the physician Ludwik Fleck argued that knowledge is first made inside a small circle of specialists, and is then changed as it moves out to a wider audience \citep{fleck_genesis_1979}.
As knowledge reaches the public, cautious claims become more direct and technical terms are replaced by everyday words.
Writing for patients therefore presents the same knowledge in a different form, adapted to a different audience.
\end{reviewread}

% === P2b — the drug leaflet: official, mandated vulgarisation ===
\begin{reviewread}
Not all of this lay writing is unofficial.
The most systematic patient-facing text of all is the drug leaflet, the \emph{notice} folded inside every box of medicine.
In the European Union, by law, every approved drug must carry one.
It must be in the official language of each market where it is sold, and written ``in clear and understandable terms for the users'' \citep{ec_directive_2001_83}.
A shared template fixes the same six headings everywhere, from \emph{what the medicine is and what it is used for} to \emph{possible side effects}.
A draft must also be tested on real patients for readability before it can be approved \citep{ema_qrd_template}.
This makes the move from specialist to public explicit.
The leaflet also shows how names vary.
The rules require the brand name to sit beside the common one, \emph{Doliprane} next to \emph{paracétamol}.
So a single substance carries several of its names side by side on one page.
\end{reviewread}

\begin{center}
\begin{tikzpicture}
\node[draw=ThesisNeutral, fill=ThesisPaper, rounded corners=3pt, line width=0.5pt,
      inner sep=10pt] {%
\begin{minipage}{0.84\textwidth}
\footnotesize\color{ThesisInk}
{\tiny\sffamily\color{ThesisNeutral} EMEA · PATIENT-FACING MEDICINE SUMMARY}\\[5pt]
\textit{Ultratard est une insuline de substitution identique à l'insuline fabriquée
par le pancréas. Le principe actif d'Ultratard, l'insuline humaine (ADNr), est
produit grâce à une méthode dite de ``technologie recombinante''.}
\end{minipage}
};
\end{tikzpicture}
\captionof{figure}{One medicine named as a brand, a treatment, and an active substance
in the EMEA split of the biomed-fr-v2 corpus.}
\label{fig:bt-emea-names}
\end{center}
\medskip

% === P3 — naming varies (Wüster foil, Cabré, Gaudin) ===
\begin{reviewread}
What changes most, for our purpose, is the naming.
When terminology first emerged as a field, the Austrian engineer Eugen Wüster argued that each concept should have one precise term, preventing ambiguity in technical communication \citep{wuster_einfuhrung_1979}.
In practice, however, the same concept often has several names, as \Cref{fig:bt-emea-names} shows for a single medicine.
María Teresa Cabré showed that the term people choose depends on the context and audience.
Rather than being an exception, this variation is a normal part of specialised language \citep{cabre_theories_2003}.
François Gaudin's socioterminology studies this variation as a social phenomenon \citep{gaudin_socioterminologie_2003}.
The heart attack of the clinical note, \emph{infarctus du myocarde}, coded \emph{I21}, is the same event the patient calls a \emph{crise cardiaque}.
One concept, many names.
\end{reviewread}

% === P3b — Temmerman: concepts are fuzzy too ===
\begin{reviewread}
The variation goes beyond words.
Medical concepts themselves can be understood in different ways.
The heart, for example, can be described as a pump, leading to terms about pressure, output, and failure.
It can also be described as an electrical system, leading to terms about rhythm, conduction, and arrhythmia.
\citet{temmerman_towards_2000} showed that such ways of understanding a concept help shape the terminology used to describe it.
Terminology must therefore account for different names and for the different views of a concept that those names express.
\end{reviewread}

% === P4 — UMLS, measured ===
\begin{reviewread}
Medical terminology systems have evolved to represent this variation.
The Unified Medical Language System brings together more than sixty medical vocabularies and links their terms to shared concepts \citep{bodenreider_unified_2004}.
These links form a many-to-many relationship: one concept can have several names, and one name can refer to several concepts.
In 2004, the UMLS contained about 2.5 million names for 900{,}000 concepts, or almost three names per concept.
It links these different names to shared concepts rather than selecting a single correct term.
\end{reviewread}

% === P5 — register, brief ===
\begin{reviewread}
Patient-facing language is itself diverse.
A forum post is personal and spontaneous, while a drug leaflet is officially written, carefully structured, and directly addresses the reader.
The CLEAR corpus makes this adaptation explicit by pairing specialised medical texts with versions written in plain language \citep{grabar_clear_2018}, as \Cref{fig:bt-clear-register} shows on a single warning.
\end{reviewread}

\begin{center}
\begin{tikzpicture}
\node[draw=ThesisNeutral, fill=ThesisPaper, rounded corners=3pt, line width=0.5pt,
      inner sep=10pt] {%
\begin{tabular}{@{}l !{\hspace{6pt}\textcolor{ThesisNeutral}{\vrule width 0.3pt}\hspace{6pt}} l@{}}
\begin{minipage}[t]{0.40\textwidth}
\footnotesize\color{ThesisInk}
{\tiny\sffamily\color{ThesisNeutral} EXPERT VERSION · FRANÇAIS}\\[5pt]
\textit{Chez les patients épileptiques ou à risque de convulsions, il est recommandé
d'évaluer le rapport bénéfice-risque du thiocolchicoside et de renforcer la
surveillance clinique. La survenue de crises convulsives impose l'arrêt du
traitement.}\\[8pt]
{\tiny\sffamily\color{ThesisNeutral} ENGLISH · TRANSLATION}\\[5pt]
\textit{For patients with epilepsy or at risk of seizures, the benefit--risk balance
of thiocolchicoside should be assessed and clinical monitoring should be increased.
If seizures occur, treatment must be stopped.}
\end{minipage}
&
\begin{minipage}[t]{0.40\textwidth}
\footnotesize\color{ThesisNeutral}
{\tiny\sffamily\color{ThesisNeutral} PATIENT VERSION · FRANÇAIS}\\[5pt]
\textit{Utiliser avec précaution en cas d'antécédents d'épilepsie ou de convulsions.
La survenue de crise convulsive impose l'arrêt du traitement.}\\[8pt]
{\tiny\sffamily\color{ThesisNeutral} ENGLISH · TRANSLATION}\\[5pt]
\textit{Use with caution if you have a history of epilepsy or seizures. If a seizure
occurs, stop treatment.}
\end{minipage}
\end{tabular}%
};
\end{tikzpicture}
\captionof{figure}{The same warning in expert and patient versions from CLEAR; French originals and English translations.}
\label{fig:bt-clear-register}
\end{center}
\medskip

\begin{reviewread}
The expert version discusses risk assessment and clinical monitoring, while the patient version gives a direct instruction.
Public medical text therefore spans several registers, each adapting medical knowledge to a different reader.
\end{reviewread}

% === P6 — synthesis + bridge (closes the part) ===
\begin{reviewread}
Biomedical text is not a single language.
The same medical knowledge changes with who writes it, who reads it, and what the text is meant to do.
Clinical notes, research articles, and patient-facing texts therefore offer different views of medicine, and only the latter two are widely available.
The challenge for the rest of this thesis is to use this public text without losing sight of the clinical language we ultimately need.
\end{reviewread}
